---
title: "Disaster Damage and Response Estimation"
author: "Théophile L. Mouton"
date: "`r Sys.Date()`"
include-after-body: "asset/footer.html"
css: "asset/style.css"
format:
  html:
    template-partials: 
      - asset/title-block.html
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    highlight-style: github
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    theme: cosmo
    fig-width: 8
    fig-height: 6
    fig-dpi: 300
    self-contained: false
execute:
  echo: true
  warning: false
  message: false
  error: false
---

# Intro

This report runs based on a config file which provides the cyclone intensity applied to each council in Vanuatu. It has 5 columns:

- National
- Province
- Area Council
- Hazard
- Intensity (number from 2 to 3)

Let's load this file to use it later in our analysis:

```{r}
config <- read.csv(here::here("Data", "Ex_hazard_areas.csv"), check.names = FALSE)
```




Another file that will be useful: a 2 columns dataframe that provides the list of councils with their Province
```{r}
council_province_lookup <- read.csv("data/council_province_lookup.csv")
```




And load the required libs
```{r, warning=FALSE, message=FALSE}
# Load required packages
library(dplyr)
library(tidyr)
library(reactable)
library(htmltools)
library(readxl)
library(here)
```





Order of councils and provinces in the final tables:

```{r}
# Define the custom order for regions
region_order <- c(
  "National", # Will be displayed as Vanuatu
  "Torba", "Torres", "Ureparapara", "Motalava", "West Vanualava", "East Vanualava", 
  "Mota", "East Gaua", "West Gaua", "Merelava",
  "Sanma", "Luganville", "North West Santo", "Big Bay Coast", "Big Bay Inland", 
  "West Santo", "South Santo 1", "South Santo 2", "East Santo", "South East Santo", 
  "Canal - Fanafo", "West Malo", "East Malo",
  "Penama", "West Ambae", "North Ambae", "East Ambae", "South Ambae", "North Maewo", 
  "South Maewo", "North Pentecost", "Central Pentecost 1", "Central Pentecost 2", 
  "South Pentecost",
  "Malampa", "North West Malekula", "North East Malekula", "Central Malekula", 
  "South West Malekula", "South East Malekula", "South Malekula", "North Ambrym", 
  "West Ambrym", "South East Ambrym", "Paama",
  "Shefa", "Port Vila", "Vermali", "Vermaul", "Varisu", "South Epi", "North Tongoa", 
  "Tongariki", "Makimae", "Nguna", "Emau", "Malorua", "North Efate", "Mele", 
  "Tanvasoko", "Ifira", "Pango", "Erakor", "Eratap", "Eton",
  "Tafea", "North Erromango", "South Erromango", "Aniwa", "North Tanna", "West Tanna", 
  "Middle Bush Tanna", "South West Tanna", "Whitesands", "South Tanna", "Futuna", "Aneityum"
)

# Provincial levels (to be bolded)
provinces <- c("National", "Torba", "Sanma", "Penama", "Malampa", "Shefa", "Tafea")
```





A function to aggregate a table with one row per council. It adds value per province and for the whole country
```{r}
# council_data must be a data frame with a column called "Region"
compute_council_aggregates <- function(council_data) {
  
  province_data <- council_data %>%
    left_join(council_province_lookup, by = c("Region" = "Council")) %>%
    group_by(Province) %>%
    summarise(across(where(is.numeric), sum, na.rm = TRUE), .groups = "drop") %>%
    rename(Region = Province)  
  
  national_data <- province_data %>%
    summarise(across(where(is.numeric), sum, na.rm = TRUE)) %>%
    mutate(Region = "National")
  
  result <- bind_rows(
    national_data,
    province_data,
    council_data
  )
  
  # Add default order and formatting columns
  result <- result %>%
    mutate(default_order = match(Region, region_order)) %>%
    arrange(default_order) %>%
    select(-default_order)

  return(result)
}

format_table <- function(data){
  result <- data %>%
    select(Region, everything()) %>% 
    mutate(
      Region = ifelse(
        Region %in% provinces, 
        paste0("<b>", Region, "</b>"), 
        Region
      )
    )
  return(result)
}
```

# Education

## Baseline: Number of Schools, Students, and Teachers

```{r}
full_data <- read.csv(here::here("Data", "1c_input_baseline.csv"), check.names = FALSE)

# === DATA WRANGLING ===
# Filter for Education in Baseline column and Area Council level data
education_data <- full_data %>%
  filter(Baseline == "Education") %>%  # Changed to filter Baseline column
  filter(Year == max(Year, na.rm = TRUE)) %>%  # Filter for the latest year
  filter(!is.na(`Area Council`)) %>%
  mutate(Region = `Area Council`)

# Reshape the data to get Schools, Students, Teachers as separate columns
education_wide <- education_data %>%
  mutate(
    education_level = tolower(Attribute),  # Keep lowercase for consistency
    measure_type = case_when(
      grepl("Number Schools", Indicator) ~ "schools",
      grepl("Students", Indicator) ~ "students",
      grepl("Teachers", Indicator) ~ "teachers",
      TRUE ~ "other"
    )
  ) %>%
  filter(measure_type %in% c("schools", "students", "teachers")) %>%
  select(Region, education_level, measure_type, Value) %>%
  pivot_wider(
    names_from = c(education_level, measure_type),
    values_from = Value,
    names_sep = "_"
  ) %>%
  mutate(across(where(is.numeric), ~replace_na(.x, 0)))

# Compute aggregates (province and national levels)
education_aggregated <- compute_council_aggregates(education_wide)

# === EXPORT TO CSV ===
write.csv(
  education_aggregated %>% select(Region, everything()),
  here::here("output", "Education_01_baseline_education.csv"),
  row.names = FALSE
)

# === PRESENTATION ===
# Format for display
formatted <- format_table(education_aggregated)

# Create the reactable with proper case labels
reactable(
  formatted,
  columns = list(
    Region = colDef(
      name = "Region", 
      minWidth = 150,
      sortable = TRUE,
      defaultSortOrder = "asc",
      html = TRUE,
      sticky = "left"
    ),
    
    # ECCE columns
    ecce_schools = colDef(name = "Schools", format = colFormat(digits = 0)),
    ecce_students = colDef(name = "Students", format = colFormat(digits = 0)),
    ecce_teachers = colDef(name = "Teachers", format = colFormat(digits = 0)),
    
    # Primary columns  
    primary_schools = colDef(name = "Schools", format = colFormat(digits = 0)),
    primary_students = colDef(name = "Students", format = colFormat(digits = 0)),
    primary_teachers = colDef(name = "Teachers", format = colFormat(digits = 0)),
    
    # Secondary columns
    secondary_schools = colDef(name = "Schools", format = colFormat(digits = 0)),
    secondary_students = colDef(name = "Students", format = colFormat(digits = 0)),
    secondary_teachers = colDef(name = "Teachers", format = colFormat(digits = 0))
  ),
  columnGroups = list(
    colGroup(name = "Region", columns = c("Region")),
    colGroup(name = "ECCE", columns = c("ecce_schools", "ecce_students", "ecce_teachers")),
    colGroup(name = "Primary", columns = c("primary_schools", "primary_students", "primary_teachers")),
    colGroup(name = "Secondary", columns = c("secondary_schools", "secondary_students", "secondary_teachers"))
  ),
  striped = TRUE,
  highlight = TRUE,
  bordered = TRUE,
  theme = reactableTheme(
    headerStyle = list(
      "&:hover" = list(background = "#eee")
    ),
    borderColor = "#ddd",
    stripedColor = "#f6f8fa"
  ),
  defaultPageSize = 15,
  showPageSizeOptions = TRUE,
  pageSizeOptions = c(10, 15, 25, 50, 75, 100),
  showSortable = TRUE
)
```





## Estimating damage: Number of damaged schools and students affected

```{r}
# === DATA WRANGLING ===
# Add cyclone strength to baseline values
education_with_config <- education_wide %>%
  left_join(config, by = c("Region" = "Area Council"))

# Load the baseline multiplying factors: 2a Input- Damage Multiplier (sheet 6)
# Read CSV while preserving column names with spaces
baseline_factors <- read.csv(
  here::here("Data", "2a Input- Damage Multiplier.csv"),
  check.names = FALSE
)

# Helper function for damage multipliers (now considers council-specific values)
get_damage_multiplier <- function(cyclone_category, education_level, measure_type, area_council) {
  intensity_col <- paste0("Intensity ", cyclone_category)
  
  multiplier <- baseline_factors %>%
    filter(
      Cluster == "Education",
      Attribute == education_level,
      grepl(measure_type, Indicator, ignore.case = TRUE),
      `Area Council` == area_council  # Add council-specific filter
    )
  
  # Check if we got any results
  if(nrow(multiplier) == 0) {
    return(0)
  }
  
  # Extract the value from the intensity column
  result <- multiplier[[intensity_col]][1]
  
  return(ifelse(is.na(result) || length(result) == 0, 0, result))
}

# Calculate damage estimates
damage_estimates <- education_with_config %>%
  filter(!is.na(Intensity)) %>%
  rowwise() %>%
  mutate(
    # ECCE - now passing Region to get council-specific multipliers
    ecce_schools_damaged = round(ecce_schools * get_damage_multiplier(Intensity, "ecce", "schools", Region)),
    ecce_students_affected = round(ecce_students * get_damage_multiplier(Intensity, "ecce", "students", Region)),
    
    # Primary
    primary_schools_damaged = round(primary_schools * get_damage_multiplier(Intensity, "primary", "schools", Region)),
    primary_students_affected = round(primary_students * get_damage_multiplier(Intensity, "primary", "students", Region)),
    
    # Secondary
    secondary_schools_damaged = round(secondary_schools * get_damage_multiplier(Intensity, "secondary", "schools", Region)),
    secondary_students_affected = round(secondary_students * get_damage_multiplier(Intensity, "secondary", "students", Region))
  ) %>%
  ungroup() %>%
  select(Region, contains("_damaged"), contains("_affected"))

# Aggregate to province and national levels
damage_estimates_full <- compute_council_aggregates(damage_estimates)

# === EXPORT TO CSV ===
write.csv(
  damage_estimates_full %>% select(Region, everything()),
  here::here("output", "Education_02_damage_estimates.csv"),
  row.names = FALSE
)
# === PRESENTATION ===
# Format for display
formatted <- format_table(damage_estimates_full)

# Create the damage estimation reactable
reactable(
  formatted,
  columns = list(
    Region = colDef(
      name = "Region", 
      minWidth = 150,
      sticky = "left",
      html = TRUE
    ),
    
    # ECCE columns
    ecce_schools_damaged = colDef(name = "Schools Damaged", format = colFormat(digits = 0)),
    ecce_students_affected = colDef(name = "Students Affected", format = colFormat(digits = 0)),
    
    # Primary columns  
    primary_schools_damaged = colDef(name = "Schools Damaged", format = colFormat(digits = 0)),
    primary_students_affected = colDef(name = "Students Affected", format = colFormat(digits = 0)),
    
    # Secondary columns
    secondary_schools_damaged = colDef(name = "Schools Damaged", format = colFormat(digits = 0)),
    secondary_students_affected = colDef(name = "Students Affected", format = colFormat(digits = 0))
  ),
  columnGroups = list(
    colGroup(name = "Region", columns = c("Region")),
    colGroup(name = "ECCE", columns = c("ecce_schools_damaged", "ecce_students_affected")),
    colGroup(name = "Primary", columns = c("primary_schools_damaged", "primary_students_affected")),
    colGroup(name = "Secondary", columns = c("secondary_schools_damaged", "secondary_students_affected"))
  ),
  striped = TRUE,
  highlight = TRUE,
  bordered = TRUE,
  theme = reactableTheme(
    headerStyle = list(
      "&:hover" = list(background = "#eee")
    ),
    borderColor = "#ddd",
    stripedColor = "#f6f8fa"
  ),
  defaultPageSize = 15,
  showPageSizeOptions = TRUE,
  pageSizeOptions = c(10, 15, 25, 50, 75, 100),
  showSortable = TRUE
)

```




## Resources needed to be sent to those affected

```{r}
# === DATA WRANGLING ===
# Load the resource response configuration "3aii Input- Resources Sent" (sheet 10)

resource_config <- read.csv(here::here("Data", "3aii Input- Resources Sent.csv"), check.names = FALSE)

# Helper function to get resource multipliers from config
get_resource_multiplier <- function(resource_type) {
  multiplier <- resource_config %>%
    filter(Cluster == "Education", Indicator == resource_type) %>%
    pull(Value)
  
  return(ifelse(length(multiplier) > 0, multiplier, 0))
}

# Configuration parameters
days_of_support <- 14  # Number of days to provide resources

# Add cyclone categories from config file to education baseline data
education_with_config_resources <- education_wide %>%
  left_join(config, by = c("Region" = "Area Council"))

# Calculate resources needed using council-specific cyclone categories
resources_needed <- education_with_config_resources %>%
  filter(!is.na(Intensity)) %>%
  rowwise() %>%
  mutate(
    # Get damage multiplier for this council's specific cyclone category
    damage_multiplier = get_damage_multiplier(Intensity, "ecce", "schools", Region),
    
    # ECCE Resources
    ecce_tents = round(ecce_schools * damage_multiplier * get_resource_multiplier("Tent"), 0),
    ecce_solar_lamps = round(ecce_schools * damage_multiplier * get_resource_multiplier("Solar Lamp"), 0),
    ecce_water = ceiling(ecce_students * damage_multiplier * get_resource_multiplier("Water") * days_of_support),
    ecce_tin_fish = round(ecce_students * damage_multiplier * get_resource_multiplier("Tin Fish") * days_of_support, 0),  # Changed to "Tin Fish"
    
    # Primary Resources
    primary_tents = round(primary_schools * damage_multiplier * get_resource_multiplier("Tent"), 0),
    primary_solar_lamps = round(primary_schools * damage_multiplier * get_resource_multiplier("Solar Lamp"), 0),
    primary_water = ceiling(primary_students * damage_multiplier * get_resource_multiplier("Water") * days_of_support),
    primary_tin_fish = round(primary_students * damage_multiplier * get_resource_multiplier("Tin Fish") * days_of_support, 0),  # Changed to "Tin Fish"
    
    # Secondary Resources
    secondary_tents = round(secondary_schools * damage_multiplier * get_resource_multiplier("Tent"), 0),
    secondary_solar_lamps = round(secondary_schools * damage_multiplier * get_resource_multiplier("Solar Lamp"), 0),
    secondary_water = ceiling(secondary_students * damage_multiplier * get_resource_multiplier("Water") * days_of_support),
    secondary_tin_fish = round(secondary_students * damage_multiplier * get_resource_multiplier("Tin Fish") * days_of_support, 0),  # Changed to "Tin Fish"
    
    # Total Resources (sum across all education levels)
    total_tents = ecce_tents + primary_tents + secondary_tents,
    total_solar_lamps = ecce_solar_lamps + primary_solar_lamps + secondary_solar_lamps,
    total_water = ecce_water + primary_water + secondary_water,
    total_tin_fish = ecce_tin_fish + primary_tin_fish + secondary_tin_fish
  ) %>%
  ungroup() %>%
  select(Region, contains("tents"), contains("solar_lamps"), contains("water"), contains("tin_fish"))

# Aggregate to province and national levels
resources_needed_full <- compute_council_aggregates(resources_needed)

# === EXPORT TO CSV ===
write.csv(
  resources_needed_full %>% select(Region, everything()),
  here::here("output", "Education_03_resources_needed.csv"),
  row.names = FALSE
)

# === PRESENTATION ===
# Format for display
formatted <- format_table(resources_needed_full)

# Create the resources needed reactable
reactable(
  formatted,
  columns = list(
    Region = colDef(
      name = "Region", 
      minWidth = 150,
      sticky = "left",
      html = TRUE
    ),
    
    # ECCE columns
    ecce_tents = colDef(name = "Tents", format = colFormat(digits = 0)),
    ecce_solar_lamps = colDef(name = "Solar Lamps", format = colFormat(digits = 0)),
    ecce_water = colDef(name = "Water (1L bottles)", format = colFormat(digits = 0)),
    ecce_tin_fish = colDef(name = "Tinned Fish", format = colFormat(digits = 0)),
    
    # Primary columns  
    primary_tents = colDef(name = "Tents", format = colFormat(digits = 0)),
    primary_solar_lamps = colDef(name = "Solar Lamps", format = colFormat(digits = 0)),
    primary_water = colDef(name = "Water (1L bottles)", format = colFormat(digits = 0)),
    primary_tin_fish = colDef(name = "Tinned Fish", format = colFormat(digits = 0)),
    
    # Secondary columns
    secondary_tents = colDef(name = "Tents", format = colFormat(digits = 0)),
    secondary_solar_lamps = colDef(name = "Solar Lamps", format = colFormat(digits = 0)),
    secondary_water = colDef(name = "Water (1L bottles)", format = colFormat(digits = 0)),
    secondary_tin_fish = colDef(name = "Tinned Fish", format = colFormat(digits = 0)),
    
    # Total columns
    total_tents = colDef(name = "Tents", format = colFormat(digits = 0)),
    total_solar_lamps = colDef(name = "Solar Lamps", format = colFormat(digits = 0)),
    total_water = colDef(name = "Water (1L bottles)", format = colFormat(digits = 0)),
    total_tin_fish = colDef(name = "Tinned Fish", format = colFormat(digits = 0))
  ),
  columnGroups = list(
    colGroup(name = "Region", columns = c("Region")),
    colGroup(name = "ECCE", columns = c("ecce_tents", "ecce_solar_lamps", "ecce_water", "ecce_tin_fish")),
    colGroup(name = "Primary", columns = c("primary_tents", "primary_solar_lamps", "primary_water", "primary_tin_fish")),
    colGroup(name = "Secondary", columns = c("secondary_tents", "secondary_solar_lamps", "secondary_water", "secondary_tin_fish")),
    colGroup(name = "Totals", columns = c("total_tents", "total_solar_lamps", "total_water", "total_tin_fish"))
  ),
  striped = TRUE,
  highlight = TRUE,
  bordered = TRUE,
  theme = reactableTheme(
    headerStyle = list(
      "&:hover" = list(background = "#eee")
    ),
    borderColor = "#ddd",
    stripedColor = "#f6f8fa"
  ),
  defaultPageSize = 15,
  showPageSizeOptions = TRUE,
  pageSizeOptions = c(10, 15, 25, 50, 75, 100),
  showSortable = TRUE
)
```




## Estimate financial damage from cyclone

```{r}

# === DATA WRANGLING ===
# Load the financial damage estimation config (4a Input- Finance Estimates, sheet 13)
# Read CSV while preserving column names with spaces
financial_config <- read.csv(
  here::here("Data", "4a Input- Finance Estimates.csv"),
  check.names = FALSE
)

# Remove empty column if exists
if("" %in% names(financial_config)) {
  financial_config <- financial_config %>% select(-1)
}

# Clean column names to remove any extra spaces
names(financial_config) <- trimws(names(financial_config))

# Helper function to get unit costs from config (now council-specific)
get_unit_cost <- function(school_type, area_council) {
  unit_cost <- financial_config %>%
    filter(
      trimws(Cluster) == "Education", 
      trimws(tolower(Attribute)) == tolower(school_type),
      grepl("Number Schools", trimws(Indicator), ignore.case = TRUE),
      `Area Council` == area_council  # Add council-specific filter
    ) %>%
    pull(Value)
  
  return(ifelse(length(unit_cost) > 0, unit_cost, 0))
}

# Add cyclone categories from config file to education baseline data
education_with_config_financial <- education_wide %>%
  left_join(config, by = c("Region" = "Area Council"))

# Calculate financial damage using council-specific cyclone categories
financial_damage <- education_with_config_financial %>%
  filter(!is.na(Intensity)) %>%
  rowwise() %>%
  mutate(
    # Get council-specific damage multipliers for each education level
    ecce_damage_multiplier = get_damage_multiplier(Intensity, "ecce", "schools", Region),
    primary_damage_multiplier = get_damage_multiplier(Intensity, "primary", "schools", Region),
    secondary_damage_multiplier = get_damage_multiplier(Intensity, "secondary", "schools", Region),
    
    # Calculate financial damage per education level (in VT) using council-specific values
    ecce_financial_damage = ecce_schools * get_unit_cost("ecce", Region) * ecce_damage_multiplier,
    primary_financial_damage = primary_schools * get_unit_cost("primary", Region) * primary_damage_multiplier,
    secondary_financial_damage = secondary_schools * get_unit_cost("secondary", Region) * secondary_damage_multiplier,
    
    # Calculate total financial damage
    total_financial_damage = ecce_financial_damage + primary_financial_damage + secondary_financial_damage
  ) %>%
  ungroup() %>%
  select(Region, contains("financial_damage"))

# Aggregate to province and national levels
financial_damage_full <- compute_council_aggregates(financial_damage)

# === EXPORT TO CSV ===
write.csv(
  financial_damage_full %>% select(Region, everything()),
  here::here("output", "Education_04_financial_damage.csv"),
  row.names = FALSE
)


# === PRESENTATION ===
# Format for display
formatted <- format_table(financial_damage_full)

# Create the financial damage reactable
reactable(
  formatted,
  columns = list(
    Region = colDef(
      name = "Region", 
      minWidth = 150,
      sticky = "left",
      html = TRUE
    ),
    
    # Financial damage columns
    ecce_financial_damage = colDef(
      name = "ECCE", 
      format = colFormat(suffix = " VT", separators = TRUE, digits = 0)
    ),
    primary_financial_damage = colDef(
      name = "Primary", 
      format = colFormat(suffix = " VT", separators = TRUE, digits = 0)
    ),
    secondary_financial_damage = colDef(
      name = "Secondary", 
      format = colFormat(suffix = " VT", separators = TRUE, digits = 0)
    ),
    total_financial_damage = colDef(
      name = "Total", 
      format = colFormat(suffix = " VT", separators = TRUE, digits = 0)
    )
  ),
  columnGroups = list(
    colGroup(name = "Region", columns = c("Region")),
    colGroup(name = "Financial Damage (VT)", columns = c("ecce_financial_damage", "primary_financial_damage", "secondary_financial_damage", "total_financial_damage"))
  ),
  striped = TRUE,
  highlight = TRUE,
  bordered = TRUE,
  theme = reactableTheme(
    headerStyle = list(
      "&:hover" = list(background = "#eee")
    ),
    borderColor = "#ddd",
    stripedColor = "#f6f8fa"
  ),
  defaultPageSize = 15,
  showPageSizeOptions = TRUE,
  pageSizeOptions = c(10, 15, 25, 50, 75, 100),
  showSortable = TRUE
)

```



# Emergency Telecommunications

## Baseline: Number of Telecommunication Towers

```{r}
# === DATA WRANGLING ===
# Filter for Emergency Telecommunications in Baseline column and Area Council level data
telecom_data <- full_data %>%
  filter(Baseline == "Emergency Telecommunications") %>%
  filter(Year == max(Year, na.rm = TRUE)) %>%
  filter(!is.na(`Area Council`)) %>%
  mutate(Region = `Area Council`)

# Reshape the data to get Digicel and Vodafone as separate columns
telecom_wide <- telecom_data %>%
  mutate(
    provider = tolower(Attribute)  # digicel or vodafone
  ) %>%
  filter(provider %in% c("digicel", "vodafone")) %>%
  select(Region, provider, Value) %>%
  # Group and sum to handle duplicates (4 councils have duplicate entries)
  group_by(Region, provider) %>%
  summarise(Value = sum(Value, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(
    names_from = provider,
    values_from = Value,
    names_prefix = "towers_"
  ) %>%
  mutate(across(where(is.numeric), ~replace_na(.x, 0)))

# Compute aggregates (province and national levels)
telecom_aggregated <- compute_council_aggregates(telecom_wide)

# === EXPORT TO CSV ===
write.csv(
  telecom_aggregated %>% select(Region, everything()),
  here::here("output", "Telecom_01_baseline_towers.csv"),
  row.names = FALSE
)

# === PRESENTATION ===
# Format for display
formatted <- format_table(telecom_aggregated)

# Create the reactable
reactable(
  formatted,
  columns = list(
    Region = colDef(
      name = "Region", 
      minWidth = 150,
      sortable = TRUE,
      defaultSortOrder = "asc",
      html = TRUE,
      sticky = "left"
    ),
    towers_digicel = colDef(
      name = "Digicel", 
      format = colFormat(digits = 0)
    ),
    towers_vodafone = colDef(
      name = "Vodafone", 
      format = colFormat(digits = 0)
    )
  ),
  columnGroups = list(
    colGroup(name = "Region", columns = c("Region")),
    colGroup(name = "Number of Towers", columns = c("towers_digicel", "towers_vodafone"))
  ),
  striped = TRUE,
  highlight = TRUE,
  bordered = TRUE,
  theme = reactableTheme(
    headerStyle = list(
      "&:hover" = list(background = "#eee")
    ),
    borderColor = "#ddd",
    stripedColor = "#f6f8fa"
  ),
  defaultPageSize = 15,
  showPageSizeOptions = TRUE,
  pageSizeOptions = c(10, 15, 25, 50, 75, 100),
  showSortable = TRUE
)
```


## Estimating Damage: Number of Damaged Towers


```{r}
# === DATA WRANGLING ===
# Add cyclone strength to baseline values
telecom_with_config <- telecom_wide %>%
  left_join(config, by = c("Region" = "Area Council"))

# Helper function for damage multipliers (council and provider-specific)
get_damage_multiplier <- function(cyclone_category, provider, area_council) {
  intensity_col <- paste0("Intensity ", cyclone_category)
  
  multiplier <- baseline_factors %>%
    filter(
      Cluster == "Emergency Telecommunications",
      Attribute == provider,
      grepl("Number Towers", Indicator, ignore.case = TRUE),
      `Area Council` == area_council
    )
  
  # Check if we got any results
  if(nrow(multiplier) == 0) {
    return(0)
  }
  
  # Extract the value from the intensity column
  result <- multiplier[[intensity_col]][1]
  
  return(ifelse(is.na(result) || length(result) == 0, 0, result))
}

# Calculate damage estimates
damage_estimates <- telecom_with_config %>%
  filter(!is.na(Intensity)) %>%
  rowwise() %>%
  mutate(
    # Digicel towers damaged
    towers_digicel_damaged = towers_digicel * get_damage_multiplier(Intensity, "digicel", Region),
    
    # Vodafone towers damaged
    towers_vodafone_damaged = towers_vodafone * get_damage_multiplier(Intensity, "vodafone", Region),
    
    # Total towers damaged
    total_towers_damaged = towers_digicel_damaged + towers_vodafone_damaged
  ) %>%
  ungroup() %>%
  select(Region, towers_digicel_damaged, towers_vodafone_damaged, total_towers_damaged)

# Aggregate to province and national levels
damage_estimates_full <- compute_council_aggregates(damage_estimates)

# === EXPORT TO CSV ===
write.csv(
  damage_estimates_full %>% select(Region, everything()),
  here::here("output", "Telecom_02_damage_estimates.csv"),
  row.names = FALSE
)

# === PRESENTATION ===
# Format for display
formatted <- format_table(damage_estimates_full)

# Create the damage estimation reactable
reactable(
  formatted,
  columns = list(
    Region = colDef(
      name = "Region", 
      minWidth = 150,
      sticky = "left",
      html = TRUE
    ),
    towers_digicel_damaged = colDef(
      name = "Digicel", 
      format = colFormat(digits = 1)
    ),
    towers_vodafone_damaged = colDef(
      name = "Vodafone", 
      format = colFormat(digits = 1)
    ),
    total_towers_damaged = colDef(
      name = "Total", 
      format = colFormat(digits = 1)
    )
  ),
  columnGroups = list(
    colGroup(name = "Region", columns = c("Region")),
    colGroup(name = "Damaged Towers", columns = c("towers_digicel_damaged", "towers_vodafone_damaged", "total_towers_damaged"))
  ),
  striped = TRUE,
  highlight = TRUE,
  bordered = TRUE,
  theme = reactableTheme(
    headerStyle = list(
      "&:hover" = list(background = "#eee")
    ),
    borderColor = "#ddd",
    stripedColor = "#f6f8fa"
  ),
  defaultPageSize = 15,
  showPageSizeOptions = TRUE,
  pageSizeOptions = c(10, 15, 25, 50, 75, 100),
  showSortable = TRUE
)

```




# Energy 


## Baseline: Household Energy Access and Usage



```{r}
# === DATA WRANGLING ===

# Get total households from Shelter baseline
total_households <- full_data %>%
  filter(Baseline == "Shelter") %>%
  filter(Year == max(Year, na.rm = TRUE)) %>%
  filter(!is.na(`Area Council`)) %>%
  filter(Indicator == "Household Type", Attribute == "number households") %>%
  mutate(Region = `Area Council`) %>%
  select(Region, Value) %>%
  rename(total_households = Value)

# Get household electricity data
electricity_data <- full_data %>%
  filter(Baseline == "Energy") %>%
  filter(Year == max(Year, na.rm = TRUE)) %>%
  filter(!is.na(`Area Council`)) %>%
  filter(Indicator == "Household Electricity") %>%
  mutate(Region = `Area Council`) %>%
  select(Region, Attribute, Value)

# Reshape electricity data
electricity_wide <- electricity_data %>%
  mutate(attribute_clean = tolower(trimws(Attribute))) %>%
  group_by(Region, attribute_clean) %>%
  summarise(Value = sum(Value, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(
    names_from = attribute_clean,
    values_from = Value,
    names_prefix = "elec_"
  ) %>%
  mutate(across(where(is.numeric), ~replace_na(.x, 0)))

# Get household cooking fuel data
cooking_data <- full_data %>%
  filter(Baseline == "Energy") %>%
  filter(Year == max(Year, na.rm = TRUE)) %>%
  filter(!is.na(`Area Council`)) %>%
  filter(Indicator == "Household Cooking Fuel") %>%
  mutate(Region = `Area Council`) %>%
  select(Region, Attribute, Value)

# Reshape cooking data
cooking_wide <- cooking_data %>%
  mutate(attribute_clean = tolower(trimws(Attribute))) %>%
  group_by(Region, attribute_clean) %>%
  summarise(Value = sum(Value, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(
    names_from = attribute_clean,
    values_from = Value,
    names_prefix = "cook_"
  ) %>%
  mutate(across(where(is.numeric), ~replace_na(.x, 0)))

# Combine all energy data
energy_wide <- total_households %>%
  left_join(electricity_wide, by = "Region") %>%
  left_join(cooking_wide, by = "Region") %>%
  mutate(across(where(is.numeric), ~replace_na(.x, 0)))

# Compute aggregates (province and national levels)
energy_aggregated <- compute_council_aggregates(energy_wide)

# === EXPORT TO CSV ===
# Reorder columns to match client's specification before export
energy_aggregated_ordered <- energy_aggregated %>%
  select(
    Region,
    total_households,
    # Household Electricity in client's order
    `elec_battery lamp`,
    elec_generator,
    `elec_main grid`,
    `elec_no access`,
    `elec_solar system`,
    # Household Cooking in client's order
    `cook_bottle gas`,
    `cook_open fire`,
    `cook_solar power`,
    cook_electricity,
    `cook_wood stove`
  )

write.csv(
  energy_aggregated_ordered,
  here::here("output", "Energy_01_baseline.csv"),
  row.names = FALSE
)

# === PRESENTATION ===
# Format for display
formatted <- format_table(energy_aggregated)

# Reorder columns to match client's specification
formatted <- formatted %>%
  select(
    Region,
    total_households,
    # Household Electricity in client's order
    `elec_battery lamp`,
    elec_generator,
    `elec_main grid`,
    `elec_no access`,
    `elec_solar system`,
    # Household Cooking in client's order
    `cook_bottle gas`,
    `cook_open fire`,
    `cook_solar power`,
    cook_electricity,
    `cook_wood stove`
  )

# Create the reactable with client's specified column order
reactable(
  formatted,
  columns = list(
    Region = colDef(
      name = "Region", 
      minWidth = 150,
      sortable = TRUE,
      defaultSortOrder = "asc",
      html = TRUE,
      sticky = "left"
    ),
    # Total Households
    total_households = colDef(name = "Total", format = colFormat(digits = 0)),
    
    # Household Electricity columns (in client's order)
    `elec_battery lamp` = colDef(name = "Battery Lamp", format = colFormat(digits = 0)),
    elec_generator = colDef(name = "Generator", format = colFormat(digits = 0)),
    `elec_main grid` = colDef(name = "Main Grid", format = colFormat(digits = 0)),
    `elec_no access` = colDef(name = "No Access", format = colFormat(digits = 0)),
    `elec_solar system` = colDef(name = "Solar System", format = colFormat(digits = 0)),
    
    # Household Cooking columns (in client's order)
    `cook_bottle gas` = colDef(name = "Bottle Gas", format = colFormat(digits = 0)),
    `cook_open fire` = colDef(name = "Open Fire", format = colFormat(digits = 0)),
    `cook_solar power` = colDef(name = "Solar Power", format = colFormat(digits = 0)),
    cook_electricity = colDef(name = "Electricity", format = colFormat(digits = 0)),
    `cook_wood stove` = colDef(name = "Wood Stove", format = colFormat(digits = 0))
  ),
  columnGroups = list(
    colGroup(name = "Total Households", columns = c("total_households")),
    colGroup(name = "Household Electricity", columns = c("elec_battery lamp", "elec_generator", "elec_main grid", "elec_no access", "elec_solar system")),
    colGroup(name = "Household Cooking", columns = c("cook_bottle gas", "cook_open fire", "cook_solar power", "cook_electricity", "cook_wood stove"))
  ),
  striped = TRUE,
  highlight = TRUE,
  bordered = TRUE,
  theme = reactableTheme(
    headerStyle = list(
      "&:hover" = list(background = "#eee")
    ),
    borderColor = "#ddd",
    stripedColor = "#f6f8fa"
  ),
  defaultPageSize = 15,
  showPageSizeOptions = TRUE,
  pageSizeOptions = c(10, 15, 25, 50, 75, 100),
  showSortable = TRUE
)

```


## Estimating damage: Households affected by electricity disruption


```{r}
# === DATA WRANGLING ===
# Add cyclone strength to baseline values
energy_with_config <- energy_wide %>%
  left_join(config, by = c("Region" = "Area Council"))

# Helper function for damage multipliers (council-specific, for both electricity and cooking)
get_energy_damage_multiplier <- function(cyclone_category, indicator_type, attribute_type, area_council) {
  intensity_col <- paste0("Intensity ", cyclone_category)
  
  multiplier <- baseline_factors %>%
    filter(
      Cluster == "Energy",
      Indicator == indicator_type,
      tolower(trimws(Attribute)) == tolower(trimws(attribute_type)),
      `Area Council` == area_council
    )
  
  # Check if we got any results
  if(nrow(multiplier) == 0) {
    return(0)
  }
  
  # Extract the value from the intensity column
  result <- multiplier[[intensity_col]][1]
  
  return(ifelse(is.na(result) || length(result) == 0, 0, result))
}

# Calculate damage estimates
damage_estimates <- energy_with_config %>%
  filter(!is.na(Intensity)) %>%
  rowwise() %>%
  mutate(
    # Total households affected (same as baseline - no damage applied)
    total_households_affected = total_households,
    
    # Damaged households for each electricity type
    `elec_battery lamp_damaged` = `elec_battery lamp` * get_energy_damage_multiplier(Intensity, "Household Electricity", "battery lamp", Region),
    elec_generator_damaged = elec_generator * get_energy_damage_multiplier(Intensity, "Household Electricity", "generator", Region),
    `elec_main grid_damaged` = `elec_main grid` * get_energy_damage_multiplier(Intensity, "Household Electricity", "main grid", Region),
    `elec_no access_damaged` = `elec_no access` * get_energy_damage_multiplier(Intensity, "Household Electricity", "no access", Region),
    `elec_solar system_damaged` = `elec_solar system` * get_energy_damage_multiplier(Intensity, "Household Electricity", "solar system", Region),
    
    # Damaged households for each cooking fuel type
    `cook_bottle gas_damaged` = `cook_bottle gas` * get_energy_damage_multiplier(Intensity, "Household Cooking Fuel", "bottle gas", Region),
    `cook_open fire_damaged` = `cook_open fire` * get_energy_damage_multiplier(Intensity, "Household Cooking Fuel", "open fire", Region),
    `cook_solar power_damaged` = `cook_solar power` * get_energy_damage_multiplier(Intensity, "Household Cooking Fuel", "solar power", Region),
    cook_electricity_damaged = cook_electricity * get_energy_damage_multiplier(Intensity, "Household Cooking Fuel", "electricity", Region),
    `cook_wood stove_damaged` = `cook_wood stove` * get_energy_damage_multiplier(Intensity, "Household Cooking Fuel", "wood stove", Region)
  ) %>%
  ungroup() %>%
  select(Region, total_households_affected, contains("_damaged"))

# Aggregate to province and national levels
damage_estimates_full <- compute_council_aggregates(damage_estimates)

# === EXPORT TO CSV ===
# Reorder columns to match client's specification before export
damage_estimates_ordered <- damage_estimates_full %>%
  select(
    Region,
    total_households_affected,
    # Household Electricity in client's order
    `elec_battery lamp_damaged`,
    elec_generator_damaged,
    `elec_main grid_damaged`,
    `elec_no access_damaged`,
    `elec_solar system_damaged`,
    # Household Cooking in client's order
    `cook_bottle gas_damaged`,
    `cook_open fire_damaged`,
    `cook_solar power_damaged`,
    cook_electricity_damaged,
    `cook_wood stove_damaged`
  )

write.csv(
  damage_estimates_ordered,
  here::here("output", "Energy_02_damage_estimates.csv"),
  row.names = FALSE
)

# === PRESENTATION ===
# Format for display
formatted <- format_table(damage_estimates_ordered)

# Create the damage estimation reactable
reactable(
  formatted,
  columns = list(
    Region = colDef(
      name = "Region", 
      minWidth = 150,
      sticky = "left",
      html = TRUE
    ),
    # Total Households
    total_households_affected = colDef(name = "Total", format = colFormat(digits = 0)),
    
    # Household Electricity columns (in client's order)
    `elec_battery lamp_damaged` = colDef(name = "Battery Lamp", format = colFormat(digits = 0)),
    elec_generator_damaged = colDef(name = "Generator", format = colFormat(digits = 0)),
    `elec_main grid_damaged` = colDef(name = "Main Grid", format = colFormat(digits = 0)),
    `elec_no access_damaged` = colDef(name = "No Access", format = colFormat(digits = 0)),
    `elec_solar system_damaged` = colDef(name = "Solar System", format = colFormat(digits = 0)),
    
    # Household Cooking columns (in client's order)
    `cook_bottle gas_damaged` = colDef(name = "Bottle Gas", format = colFormat(digits = 0)),
    `cook_open fire_damaged` = colDef(name = "Open Fire", format = colFormat(digits = 0)),
    `cook_solar power_damaged` = colDef(name = "Solar Power", format = colFormat(digits = 0)),
    cook_electricity_damaged = colDef(name = "Electricity", format = colFormat(digits = 0)),
    `cook_wood stove_damaged` = colDef(name = "Wood Stove", format = colFormat(digits = 0))
  ),
  columnGroups = list(
    colGroup(name = "Total Households", columns = c("total_households_affected")),
    colGroup(name = "Household Electricity Affected", columns = c("elec_battery lamp_damaged", "elec_generator_damaged", "elec_main grid_damaged", "elec_no access_damaged", "elec_solar system_damaged")),
    colGroup(name = "Household Cooking Affected", columns = c("cook_bottle gas_damaged", "cook_open fire_damaged", "cook_solar power_damaged", "cook_electricity_damaged", "cook_wood stove_damaged"))
  ),
  striped = TRUE,
  highlight = TRUE,
  bordered = TRUE,
  theme = reactableTheme(
    headerStyle = list(
      "&:hover" = list(background = "#eee")
    ),
    borderColor = "#ddd",
    stripedColor = "#f6f8fa"
  ),
  defaultPageSize = 15,
  showPageSizeOptions = TRUE,
  pageSizeOptions = c(10, 15, 25, 50, 75, 100),
  showSortable = TRUE
)
```



## Immediate Response Resources 


```{r}
# === DATA WRANGLING ===

# First, we need to get infrastructure baseline data and calculate damage
energy_infra_data <- full_data %>%
  filter(Baseline == "Energy") %>%
  filter(Year == max(Year, na.rm = TRUE)) %>%
  filter(!is.na(`Area Council`)) %>%
  filter(Indicator == "Energy Infrastructure") %>%
  mutate(Region = `Area Council`) %>%
  select(Region, Attribute, Value)

# Reshape infrastructure data
energy_infra_wide <- energy_infra_data %>%
  mutate(attribute_clean = tolower(trimws(Attribute))) %>%
  group_by(Region, attribute_clean) %>%
  summarise(Value = sum(Value, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(
    names_from = attribute_clean,
    values_from = Value,
    names_prefix = "infra_"
  ) %>%
  mutate(across(where(is.numeric), ~replace_na(.x, 0)))

# Add cyclone config to infrastructure data
energy_infra_with_config <- energy_infra_wide %>%
  left_join(config, by = c("Region" = "Area Council"))

# Helper function for infrastructure damage multipliers
get_infra_damage_multiplier <- function(cyclone_category, attribute_type, area_council) {
  intensity_col <- paste0("Intensity ", cyclone_category)
  
  multiplier <- baseline_factors %>%
    filter(
      Cluster == "Energy",
      Indicator == "Energy Infrastructure",
      tolower(trimws(Attribute)) == tolower(trimws(attribute_type)),
      `Area Council` == area_council
    )
  
  if(nrow(multiplier) == 0) {
    return(0)
  }
  
  result <- multiplier[[intensity_col]][1]
  return(ifelse(is.na(result) || length(result) == 0, 0, result))
}

# Calculate damaged infrastructure
infra_damage <- energy_infra_with_config %>%
  filter(!is.na(Intensity)) %>%
  rowwise() %>%
  mutate(
    # Calculate damaged poles
    `infra_low voltage poles_damaged` = `infra_low voltage poles` * get_infra_damage_multiplier(Intensity, "low voltage poles", Region),
    `infra_high voltage poles_damaged` = `infra_high voltage poles` * get_infra_damage_multiplier(Intensity, "high voltage poles", Region),
    
    # Total damaged poles for power line cable calculation
    total_poles_damaged = `infra_low voltage poles_damaged` + `infra_high voltage poles_damaged`
  ) %>%
  ungroup() %>%
  select(Region, contains("_damaged"), total_poles_damaged)

# Now calculate resources needed
# Join household damage data with infrastructure damage data
resources_base <- damage_estimates %>%
  left_join(infra_damage, by = "Region")

# Calculate total electricity households affected
resources_needed <- resources_base %>%
  mutate(
    # Total households with electricity affected (for solar lamps)
    total_elec_affected = `elec_battery lamp_damaged` + elec_generator_damaged + 
                          `elec_main grid_damaged` + `elec_no access_damaged` + 
                          `elec_solar system_damaged`,
    
    # Resource calculations based on specifications
    solar_lamp = round(1 * total_elec_affected, 0),
    low_voltage_poles_support = round(0.2 * `infra_low voltage poles_damaged`, 0),
    high_voltage_pole_support = round(0.2 * `infra_high voltage poles_damaged`, 0),
    power_line_cable = round(100 * total_poles_damaged, 0)
  ) %>%
  select(Region, solar_lamp, low_voltage_poles_support, high_voltage_pole_support, power_line_cable)

# Aggregate to province and national levels
resources_needed_full <- compute_council_aggregates(resources_needed)

# === EXPORT TO CSV ===
write.csv(
  resources_needed_full,
  here::here("output", "Energy_03_resources_needed.csv"),
  row.names = FALSE
)

# === PRESENTATION ===
# Format for display
formatted <- format_table(resources_needed_full)

# Create the resources needed reactable
reactable(
  formatted,
  columns = list(
    Region = colDef(
      name = "Region", 
      minWidth = 150,
      sticky = "left",
      html = TRUE
    ),
    solar_lamp = colDef(
      name = "Solar Lamp", 
      format = colFormat(digits = 0)
    ),
    low_voltage_poles_support = colDef(
      name = "Low Voltage Poles Support", 
      format = colFormat(digits = 0)
    ),
    high_voltage_pole_support = colDef(
      name = "High Voltage Pole Support", 
      format = colFormat(digits = 0)
    ),
    power_line_cable = colDef(
      name = "Power Line Cable (metres)", 
      format = colFormat(digits = 0)
    )
  ),
  columnGroups = list(
    colGroup(name = "Region", columns = c("Region")),
    colGroup(name = "Household Electricity", columns = c("solar_lamp")),
    colGroup(name = "Non-renewable", columns = c("low_voltage_poles_support", "high_voltage_pole_support", "power_line_cable"))
  ),
  striped = TRUE,
  highlight = TRUE,
  bordered = TRUE,
  theme = reactableTheme(
    headerStyle = list(
      "&:hover" = list(background = "#eee")
    ),
    borderColor = "#ddd",
    stripedColor = "#f6f8fa"
  ),
  defaultPageSize = 15,
  showPageSizeOptions = TRUE,
  pageSizeOptions = c(10, 15, 25, 50, 75, 100),
  showSortable = TRUE
)

```


## Estimated Financial Damage

### Table 1: Household Electricity Financial Damage

```{r}
# Helper function to get unit costs for household electricity
get_household_elec_unit_cost <- function(electricity_type, area_council) {
  unit_cost <- financial_config %>%
    filter(
      Cluster == "Energy",
      Indicator == "Household Electricity",
      tolower(trimws(Attribute)) == tolower(trimws(electricity_type)),
      `Area Council` == area_council
    ) %>%
    pull(Value)
  
  return(ifelse(length(unit_cost) > 0, unit_cost, 0))
}

# Add cyclone categories from config file
energy_damage_with_config <- damage_estimates %>%
  left_join(config, by = c("Region" = "Area Council"))

# Calculate household electricity financial damage
household_elec_financial <- energy_damage_with_config %>%
  filter(!is.na(Intensity)) %>%
  rowwise() %>%
  mutate(
    # Financial damage for each electricity type (in VT)
    battery_lamp_damage = `elec_battery lamp_damaged` * get_household_elec_unit_cost("battery lamp", Region),
    generator_damage = elec_generator_damaged * get_household_elec_unit_cost("generator", Region),
    main_grid_damage = `elec_main grid_damaged` * get_household_elec_unit_cost("electricity poles", Region),  # Using "electricity poles" as proxy for main grid
    
    # Total household electricity financial damage
    total_household_elec_damage = battery_lamp_damage + generator_damage + main_grid_damage
  ) %>%
  ungroup() %>%
  select(Region, battery_lamp_damage, generator_damage, main_grid_damage, total_household_elec_damage)

# Aggregate to province and national levels
household_elec_financial_full <- compute_council_aggregates(household_elec_financial)

# === EXPORT TABLE 1 TO CSV ===
write.csv(
  household_elec_financial_full,
  here::here("output", "Energy_04a_financial_damage_household_electricity.csv"),
  row.names = FALSE
)

# === PRESENTATION TABLE 1 ===
formatted1 <- format_table(household_elec_financial_full)

reactable(
  formatted1,
  columns = list(
    Region = colDef(
      name = "Region", 
      minWidth = 150,
      sticky = "left",
      html = TRUE
    ),
    battery_lamp_damage = colDef(
      name = "Battery Lamp",
      format = colFormat(suffix = " VT", separators = TRUE, digits = 0)
    ),
    generator_damage = colDef(
      name = "Generator",
      format = colFormat(suffix = " VT", separators = TRUE, digits = 0)
    ),
    main_grid_damage = colDef(
      name = "Main Grid",
      format = colFormat(suffix = " VT", separators = TRUE, digits = 0)
    ),
    total_household_elec_damage = colDef(
      name = "Total Value",
      format = colFormat(suffix = " VT", separators = TRUE, digits = 0)
    )
  ),
  columnGroups = list(
    colGroup(name = "Region", columns = c("Region")),
    colGroup(name = "Household Electricity", columns = c("battery_lamp_damage", "generator_damage", "main_grid_damage", "total_household_elec_damage"))
  ),
  striped = TRUE,
  highlight = TRUE,
  bordered = TRUE,
  theme = reactableTheme(
    headerStyle = list(
      "&:hover" = list(background = "#eee")
    ),
    borderColor = "#ddd",
    stripedColor = "#f6f8fa"
  ),
  defaultPageSize = 15,
  showPageSizeOptions = TRUE,
  pageSizeOptions = c(10, 15, 25, 50, 75, 100),
  showSortable = TRUE
)
```


### Table 2: Energy Infrastructure Financial Damage


```{r}
# Helper function to get unit costs for infrastructure
get_infrastructure_unit_cost <- function(infrastructure_type, area_council) {
  unit_cost <- financial_config %>%
    filter(
      Cluster == "Energy",
      Indicator == "Energy Infrastructure",
      tolower(trimws(Attribute)) == tolower(trimws(infrastructure_type)),
      `Area Council` == area_council
    ) %>%
    pull(Value)
  
  return(ifelse(length(unit_cost) > 0, unit_cost, 0))
}

# Join infrastructure baseline with config
energy_infra_with_config <- energy_infra_wide %>%
  left_join(config, by = c("Region" = "Area Council"))

# Calculate infrastructure financial damage
infrastructure_financial <- energy_infra_with_config %>%
  filter(!is.na(Intensity)) %>%
  rowwise() %>%
  mutate(
    # Calculate damaged infrastructure units and their financial cost
    hydro_damaged = if("infra_hydro" %in% names(.)) infra_hydro * get_infra_damage_multiplier(Intensity, "hydro", Region) else 0,
    hydro_financial = hydro_damaged * get_infrastructure_unit_cost("hydro", Region),
    
    solar_damaged = if("infra_solar" %in% names(.)) infra_solar * get_infra_damage_multiplier(Intensity, "solar", Region) else 0,
    solar_financial = solar_damaged * get_infrastructure_unit_cost("solar", Region),
    
    pole_support_damaged = if("infra_electrcity pole support" %in% names(.)) `infra_electrcity pole support` * get_infra_damage_multiplier(Intensity, "electrcity pole support", Region) else 0,
    pole_support_financial = pole_support_damaged * get_infrastructure_unit_cost("electrcity pole support", Region),
    
    terminals_damaged = if("infra_electrical terminals" %in% names(.)) `infra_electrical terminals` * get_infra_damage_multiplier(Intensity, "electrical terminals", Region) else 0,
    terminals_financial = terminals_damaged * get_infrastructure_unit_cost("electrical terminals", Region),
    
    hv_poles_damaged = if("infra_high voltage poles" %in% names(.)) `infra_high voltage poles` * get_infra_damage_multiplier(Intensity, "high voltage poles", Region) else 0,
    hv_poles_financial = hv_poles_damaged * get_infrastructure_unit_cost("high voltage poles", Region),
    
    hv_support_damaged = if("infra_high voltage support poles" %in% names(.)) `infra_high voltage support poles` * get_infra_damage_multiplier(Intensity, "high voltage support poles", Region) else 0,
    hv_support_financial = hv_support_damaged * get_infrastructure_unit_cost("high voltage support poles", Region),
    
    hv_transformer_damaged = if("infra_high voltage transformer substation" %in% names(.)) `infra_high voltage transformer substation` * get_infra_damage_multiplier(Intensity, "high voltage transformer substation", Region) else 0,
    hv_transformer_financial = hv_transformer_damaged * get_infrastructure_unit_cost("high voltage transformer substation", Region),
    
    lv_poles_damaged = if("infra_low voltage poles" %in% names(.)) `infra_low voltage poles` * get_infra_damage_multiplier(Intensity, "low voltage poles", Region) else 0,
    lv_poles_financial = lv_poles_damaged * get_infrastructure_unit_cost("low voltage poles", Region),
    
    lv_distribution_damaged = if("infra_low voltage street distrbution boxes" %in% names(.)) `infra_low voltage street distrbution boxes` * get_infra_damage_multiplier(Intensity, "low voltage street distrbution boxes", Region) else 0,
    lv_distribution_financial = lv_distribution_damaged * get_infrastructure_unit_cost("low voltage street distrbution boxes", Region),
    
    # Total infrastructure financial damage
    total_infrastructure_damage = hydro_financial + solar_financial + pole_support_financial + 
                                  terminals_financial + hv_poles_financial + hv_support_financial +
                                  hv_transformer_financial + lv_poles_financial + lv_distribution_financial
  ) %>%
  ungroup() %>%
  select(Region, hydro_financial, solar_financial, pole_support_financial, terminals_financial,
         hv_poles_financial, hv_support_financial, hv_transformer_financial, 
         lv_poles_financial, lv_distribution_financial, total_infrastructure_damage)

# Aggregate to province and national levels
infrastructure_financial_full <- compute_council_aggregates(infrastructure_financial)

# === EXPORT TABLE 2 TO CSV ===
write.csv(
  infrastructure_financial_full,
  here::here("output", "Energy_04b_financial_damage_infrastructure.csv"),
  row.names = FALSE
)

# === PRESENTATION TABLE 2 ===
formatted2 <- format_table(infrastructure_financial_full)

reactable(
  formatted2,
  columns = list(
    Region = colDef(
      name = "Region", 
      minWidth = 150,
      sticky = "left",
      html = TRUE
    ),
    hydro_financial = colDef(name = "Hydro", format = colFormat(suffix = " VT", separators = TRUE, digits = 0)),
    solar_financial = colDef(name = "Solar", format = colFormat(suffix = " VT", separators = TRUE, digits = 0)),
    pole_support_financial = colDef(name = "Electricity Pole Support", format = colFormat(suffix = " VT", separators = TRUE, digits = 0)),
    terminals_financial = colDef(name = "Electrical Terminals", format = colFormat(suffix = " VT", separators = TRUE, digits = 0)),
    hv_poles_financial = colDef(name = "High Voltage Poles", format = colFormat(suffix = " VT", separators = TRUE, digits = 0)),
    hv_support_financial = colDef(name = "High Voltage Support Poles", format = colFormat(suffix = " VT", separators = TRUE, digits = 0)),
    hv_transformer_financial = colDef(name = "HV Transformer Substation", format = colFormat(suffix = " VT", separators = TRUE, digits = 0)),
    lv_poles_financial = colDef(name = "Low Voltage Poles", format = colFormat(suffix = " VT", separators = TRUE, digits = 0)),
    lv_distribution_financial = colDef(name = "LV Street Distribution Boxes", format = colFormat(suffix = " VT", separators = TRUE, digits = 0)),
    total_infrastructure_damage = colDef(name = "Total Value", format = colFormat(suffix = " VT", separators = TRUE, digits = 0))
  ),
  columnGroups = list(
    colGroup(name = "Region", columns = c("Region")),
    colGroup(name = "Renewable", columns = c("hydro_financial", "solar_financial")),
    colGroup(name = "Non-renewable", columns = c("pole_support_financial", "terminals_financial", 
                                                   "hv_poles_financial", "hv_support_financial",
                                                   "hv_transformer_financial", "lv_poles_financial",
                                                   "lv_distribution_financial")),
    colGroup(name = "Total", columns = c("total_infrastructure_damage"))
  ),
  striped = TRUE,
  highlight = TRUE,
  bordered = TRUE,
  theme = reactableTheme(
    headerStyle = list(
      "&:hover" = list(background = "#eee")
    ),
    borderColor = "#ddd",
    stripedColor = "#f6f8fa"
  ),
  defaultPageSize = 15,
  showPageSizeOptions = TRUE,
  pageSizeOptions = c(10, 15, 25, 50, 75, 100),
  showSortable = TRUE
)
```



# Food security 

## Baseline: Food Security

### Table 1: Staple crops

```{r}
# === DATA WRANGLING ===

# Get total households from Shelter baseline (for both tables)
total_households <- full_data %>%
  filter(Baseline == "Shelter") %>%
  filter(Year == max(Year, na.rm = TRUE)) %>%
  filter(!is.na(`Area Council`)) %>%
  filter(Indicator == "Household Type", Attribute == "number households") %>%
  mutate(Region = `Area Council`) %>%
  select(Region, Value) %>%
  rename(total_households = Value)

# === TABLE 1: STAPLE CROPS ===

# Get staple crop data
staple_crop_data <- full_data %>%
  filter(Baseline == "Food Security") %>%
  filter(Year == max(Year, na.rm = TRUE)) %>%
  filter(!is.na(`Area Council`)) %>%
  filter(Indicator %in% c("Stable Crop Households", "Staple Crop Production")) %>%
  mutate(Region = `Area Council`) %>%
  select(Region, Indicator, Attribute, Value)

# Reshape staple crop data
staple_crop_wide <- staple_crop_data %>%
  mutate(
    crop = tolower(trimws(Attribute)),
    metric = case_when(
      grepl("Households", Indicator) ~ "households",
      grepl("Production", Indicator) ~ "production",
      TRUE ~ "other"
    )
  ) %>%
  group_by(Region, crop, metric) %>%
  summarise(Value = sum(Value, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(
    names_from = c(crop, metric),
    values_from = Value,
    names_sep = "_"
  ) %>%
  mutate(across(where(is.numeric), ~replace_na(.x, 0)))

# Combine with total households
staple_crop_full <- total_households %>%
  left_join(staple_crop_wide, by = "Region") %>%
  mutate(across(where(is.numeric), ~replace_na(.x, 0)))

# Compute aggregates (province and national levels)
staple_crop_aggregated <- compute_council_aggregates(staple_crop_full)

# Reorder columns according to client's specification
staple_crop_ordered <- staple_crop_aggregated %>%
  select(
    Region,
    total_households,
    # Island Cabbage
    `island cabbage_households`,
    `island cabbage_production`,
    # Banana
    banana_households,
    banana_production,
    # Taro
    taro_households,
    taro_production,
    # Kumala
    kumala_households,
    kumala_production,
    # Manioc
    manioc_households,
    manioc_production,
    # Yam
    yam_households,
    yam_production
  )

# === EXPORT TABLE 1 TO CSV ===
write.csv(
  staple_crop_ordered,
  here::here("output", "FoodSecurity_01a_baseline_staple_crops.csv"),
  row.names = FALSE
)

# === PRESENTATION TABLE 1 ===
formatted1 <- format_table(staple_crop_ordered)

reactable(
  formatted1,
  columns = list(
    Region = colDef(
      name = "Region", 
      minWidth = 150,
      sticky = "left",
      html = TRUE
    ),
    total_households = colDef(name = "Total", format = colFormat(digits = 0)),
    
    # Island Cabbage
    `island cabbage_households` = colDef(name = "Households", format = colFormat(digits = 0)),
    `island cabbage_production` = colDef(name = "Production", format = colFormat(digits = 0)),
    
    # Banana
    banana_households = colDef(name = "Households", format = colFormat(digits = 0)),
    banana_production = colDef(name = "Production", format = colFormat(digits = 0)),
    
    # Taro
    taro_households = colDef(name = "Households", format = colFormat(digits = 0)),
    taro_production = colDef(name = "Production", format = colFormat(digits = 0)),
    
    # Kumala
    kumala_households = colDef(name = "Households", format = colFormat(digits = 0)),
    kumala_production = colDef(name = "Production", format = colFormat(digits = 0)),
    
    # Manioc
    manioc_households = colDef(name = "Households", format = colFormat(digits = 0)),
    manioc_production = colDef(name = "Production", format = colFormat(digits = 0)),
    
    # Yam
    yam_households = colDef(name = "Households", format = colFormat(digits = 0)),
    yam_production = colDef(name = "Production", format = colFormat(digits = 0))
  ),
  columnGroups = list(
    colGroup(name = "Total Households", columns = c("total_households")),
    colGroup(name = "Island Cabbage", columns = c("island cabbage_households", "island cabbage_production")),
    colGroup(name = "Banana", columns = c("banana_households", "banana_production")),
    colGroup(name = "Taro", columns = c("taro_households", "taro_production")),
    colGroup(name = "Kumala", columns = c("kumala_households", "kumala_production")),
    colGroup(name = "Manioc", columns = c("manioc_households", "manioc_production")),
    colGroup(name = "Yam", columns = c("yam_households", "yam_production"))
  ),
  striped = TRUE,
  highlight = TRUE,
  bordered = TRUE,
  theme = reactableTheme(
    headerStyle = list(
      "&:hover" = list(background = "#eee")
    ),
    borderColor = "#ddd",
    stripedColor = "#f6f8fa"
  ),
  defaultPageSize = 15,
  showPageSizeOptions = TRUE,
  pageSizeOptions = c(10, 15, 25, 50, 75, 100),
  showSortable = TRUE
)

```

### Table 2: Cash Crops

```{r}
# Get cash crop data
cash_crop_data <- full_data %>%
  filter(Baseline == "Food Security") %>%
  filter(Year == max(Year, na.rm = TRUE)) %>%
  filter(!is.na(`Area Council`)) %>%
  filter(Indicator %in% c("Cash Crop Households", "Cash Crop Production")) %>%
  mutate(Region = `Area Council`) %>%
  select(Region, Indicator, Attribute, Value)

# Reshape cash crop data
cash_crop_wide <- cash_crop_data %>%
  mutate(
    crop = tolower(trimws(Attribute)),
    metric = case_when(
      grepl("Households", Indicator) ~ "households",
      grepl("Production", Indicator) ~ "production",
      TRUE ~ "other"
    )
  ) %>%
  group_by(Region, crop, metric) %>%
  summarise(Value = sum(Value, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(
    names_from = c(crop, metric),
    values_from = Value,
    names_sep = "_"
  ) %>%
  mutate(across(where(is.numeric), ~replace_na(.x, 0)))

# Combine with total households
cash_crop_full <- total_households %>%
  left_join(cash_crop_wide, by = "Region") %>%
  mutate(across(where(is.numeric), ~replace_na(.x, 0)))

# Compute aggregates (province and national levels)
cash_crop_aggregated <- compute_council_aggregates(cash_crop_full)

# Reorder columns according to client's specification
cash_crop_ordered <- cash_crop_aggregated %>%
  select(
    Region,
    total_households,
    # Kava
    kava_households,
    kava_production,
    # Coconut
    coconut_households,
    coconut_production,
    # Cocoa
    cocoa_households,
    cocoa_production,
    # Coffee
    coffee_households,
    coffee_production,
    # Vanilla
    vanilla_households,
    vanilla_production,
    # Tahitian Lime
    `tahitian lime_households`,
    `tahitian lime_production`,
    # Pepper
    pepper_households,
    pepper_production,
    # Noni
    noni_households,
    noni_production
  )

# === EXPORT TABLE 2 TO CSV ===
write.csv(
  cash_crop_ordered,
  here::here("output", "FoodSecurity_01b_baseline_cash_crops.csv"),
  row.names = FALSE
)

# === PRESENTATION TABLE 2 ===
formatted2 <- format_table(cash_crop_ordered)

reactable(
  formatted2,
  columns = list(
    Region = colDef(
      name = "Region", 
      minWidth = 150,
      sticky = "left",
      html = TRUE
    ),
    total_households = colDef(name = "Total", format = colFormat(digits = 0)),
    
    # Kava
    kava_households = colDef(name = "Households", format = colFormat(digits = 0)),
    kava_production = colDef(name = "Production", format = colFormat(digits = 0)),
    
    # Coconut
    coconut_households = colDef(name = "Households", format = colFormat(digits = 0)),
    coconut_production = colDef(name = "Production", format = colFormat(digits = 0)),
    
    # Cocoa
    cocoa_households = colDef(name = "Households", format = colFormat(digits = 0)),
    cocoa_production = colDef(name = "Production", format = colFormat(digits = 0)),
    
    # Coffee
    coffee_households = colDef(name = "Households", format = colFormat(digits = 0)),
    coffee_production = colDef(name = "Production", format = colFormat(digits = 0)),
    
    # Vanilla
    vanilla_households = colDef(name = "Households", format = colFormat(digits = 0)),
    vanilla_production = colDef(name = "Production", format = colFormat(digits = 0)),
    
    # Tahitian Lime
    `tahitian lime_households` = colDef(name = "Households", format = colFormat(digits = 0)),
    `tahitian lime_production` = colDef(name = "Production", format = colFormat(digits = 0)),
    
    # Pepper
    pepper_households = colDef(name = "Households", format = colFormat(digits = 0)),
    pepper_production = colDef(name = "Production", format = colFormat(digits = 0)),
    
    # Noni
    noni_households = colDef(name = "Households", format = colFormat(digits = 0)),
    noni_production = colDef(name = "Production", format = colFormat(digits = 0))
  ),
  columnGroups = list(
    colGroup(name = "Total Households", columns = c("total_households")),
    colGroup(name = "Kava", columns = c("kava_households", "kava_production")),
    colGroup(name = "Coconut", columns = c("coconut_households", "coconut_production")),
    colGroup(name = "Cocoa", columns = c("cocoa_households", "cocoa_production")),
    colGroup(name = "Coffee", columns = c("coffee_households", "coffee_production")),
    colGroup(name = "Vanilla", columns = c("vanilla_households", "vanilla_production")),
    colGroup(name = "Tahitian Lime", columns = c("tahitian lime_households", "tahitian lime_production")),
    colGroup(name = "Pepper", columns = c("pepper_households", "pepper_production")),
    colGroup(name = "Noni", columns = c("noni_households", "noni_production"))
  ),
  striped = TRUE,
  highlight = TRUE,
  bordered = TRUE,
  theme = reactableTheme(
    headerStyle = list(
      "&:hover" = list(background = "#eee")
    ),
    borderColor = "#ddd",
    stripedColor = "#f6f8fa"
  ),
  defaultPageSize = 15,
  showPageSizeOptions = TRUE,
  pageSizeOptions = c(10, 15, 25, 50, 75, 100),
  showSortable = TRUE
)
```


## Estimated Hazard Damage 

### Table 1: Staple Crops Damage

```{r}
# Helper function for food security damage multipliers
get_food_security_damage_multiplier <- function(cyclone_category, indicator_type, crop_type, area_council) {
  intensity_col <- paste0("Intensity ", cyclone_category)
  
  multiplier <- baseline_factors %>%
    filter(
      Cluster == "Food Security",
      Indicator == indicator_type,
      tolower(trimws(Attribute)) == tolower(trimws(crop_type)),
      `Area Council` == area_council
    )
  
  if(nrow(multiplier) == 0) {
    return(0)
  }
  
  result <- multiplier[[intensity_col]][1]
  return(ifelse(is.na(result) || length(result) == 0, 0, result))
}

# Add cyclone categories from config file
staple_crop_with_config <- staple_crop_full %>%
  left_join(config, by = c("Region" = "Area Council"))

# Calculate staple crop damage estimates
staple_damage <- staple_crop_with_config %>%
  filter(!is.na(Intensity)) %>%
  rowwise() %>%
  mutate(
    # Total households affected (same as baseline)
    total_households_affected = total_households,
    
    # Island Cabbage
    `island cabbage_households_damaged` = `island cabbage_households` * get_food_security_damage_multiplier(Intensity, "Stable Crop Households", "island cabbage", Region),
    `island cabbage_production_damaged` = `island cabbage_production` * get_food_security_damage_multiplier(Intensity, "Stable Crop Production", "island cabbage", Region),
    
    # Banana
    banana_households_damaged = banana_households * get_food_security_damage_multiplier(Intensity, "Stable Crop Households", "banana", Region),
    banana_production_damaged = banana_production * get_food_security_damage_multiplier(Intensity, "Stable Crop Production", "banana", Region),
    
    # Taro
    taro_households_damaged = taro_households * get_food_security_damage_multiplier(Intensity, "Stable Crop Households", "taro", Region),
    taro_production_damaged = taro_production * get_food_security_damage_multiplier(Intensity, "Stable Crop Production", "taro", Region),
    
    # Kumala
    kumala_households_damaged = kumala_households * get_food_security_damage_multiplier(Intensity, "Stable Crop Households", "kumala", Region),
    kumala_production_damaged = kumala_production * get_food_security_damage_multiplier(Intensity, "Stable Crop Production", "kumala", Region),
    
    # Manioc
    manioc_households_damaged = manioc_households * get_food_security_damage_multiplier(Intensity, "Stable Crop Households", "manioc", Region),
    manioc_production_damaged = manioc_production * get_food_security_damage_multiplier(Intensity, "Stable Crop Production", "manioc", Region),
    
    # Yam
    yam_households_damaged = yam_households * get_food_security_damage_multiplier(Intensity, "Stable Crop Households", "yam", Region),
    yam_production_damaged = yam_production * get_food_security_damage_multiplier(Intensity, "Stable Crop Production", "yam", Region)
  ) %>%
  ungroup() %>%
  select(Region, total_households_affected, contains("_damaged"))

# Aggregate to province and national levels
staple_damage_full <- compute_council_aggregates(staple_damage)

# Reorder columns according to client's specification
staple_damage_ordered <- staple_damage_full %>%
  select(
    Region,
    total_households_affected,
    # Island Cabbage
    `island cabbage_households_damaged`,
    `island cabbage_production_damaged`,
    # Banana
    banana_households_damaged,
    banana_production_damaged,
    # Taro
    taro_households_damaged,
    taro_production_damaged,
    # Kumala
    kumala_households_damaged,
    kumala_production_damaged,
    # Manioc
    manioc_households_damaged,
    manioc_production_damaged,
    # Yam
    yam_households_damaged,
    yam_production_damaged
  )

# === EXPORT TABLE 1 TO CSV ===
write.csv(
  staple_damage_ordered,
  here::here("output", "FoodSecurity_02a_damage_staple_crops.csv"),
  row.names = FALSE
)

# === PRESENTATION TABLE 1 ===
formatted1 <- format_table(staple_damage_ordered)

reactable(
  formatted1,
  columns = list(
    Region = colDef(
      name = "Region", 
      minWidth = 150,
      sticky = "left",
      html = TRUE
    ),
    total_households_affected = colDef(name = "Total", format = colFormat(digits = 0)),
    
    # Island Cabbage
    `island cabbage_households_damaged` = colDef(name = "Households", format = colFormat(digits = 0)),
    `island cabbage_production_damaged` = colDef(name = "Production", format = colFormat(digits = 0)),
    
    # Banana
    banana_households_damaged = colDef(name = "Households", format = colFormat(digits = 0)),
    banana_production_damaged = colDef(name = "Production", format = colFormat(digits = 0)),
    
    # Taro
    taro_households_damaged = colDef(name = "Households", format = colFormat(digits = 0)),
    taro_production_damaged = colDef(name = "Production", format = colFormat(digits = 0)),
    
    # Kumala
    kumala_households_damaged = colDef(name = "Households", format = colFormat(digits = 0)),
    kumala_production_damaged = colDef(name = "Production", format = colFormat(digits = 0)),
    
    # Manioc
    manioc_households_damaged = colDef(name = "Households", format = colFormat(digits = 0)),
    manioc_production_damaged = colDef(name = "Production", format = colFormat(digits = 0)),
    
    # Yam
    yam_households_damaged = colDef(name = "Households", format = colFormat(digits = 0)),
    yam_production_damaged = colDef(name = "Production", format = colFormat(digits = 0))
  ),
  columnGroups = list(
    colGroup(name = "Total Households", columns = c("total_households_affected")),
    colGroup(name = "Island Cabbage", columns = c("island cabbage_households_damaged", "island cabbage_production_damaged")),
    colGroup(name = "Banana", columns = c("banana_households_damaged", "banana_production_damaged")),
    colGroup(name = "Taro", columns = c("taro_households_damaged", "taro_production_damaged")),
    colGroup(name = "Kumala", columns = c("kumala_households_damaged", "kumala_production_damaged")),
    colGroup(name = "Manioc", columns = c("manioc_households_damaged", "manioc_production_damaged")),
    colGroup(name = "Yam", columns = c("yam_households_damaged", "yam_production_damaged"))
  ),
  striped = TRUE,
  highlight = TRUE,
  bordered = TRUE,
  theme = reactableTheme(
    headerStyle = list(
      "&:hover" = list(background = "#eee")
    ),
    borderColor = "#ddd",
    stripedColor = "#f6f8fa"
  ),
  defaultPageSize = 15,
  showPageSizeOptions = TRUE,
  pageSizeOptions = c(10, 15, 25, 50, 75, 100),
  showSortable = TRUE
)
```

### Table 2: Cash Crops Damage

```{r}
# Add cyclone categories from config file
cash_crop_with_config <- cash_crop_full %>%
  left_join(config, by = c("Region" = "Area Council"))

# Calculate cash crop damage estimates
cash_damage <- cash_crop_with_config %>%
  filter(!is.na(Intensity)) %>%
  rowwise() %>%
  mutate(
    # Total households affected (same as baseline)
    total_households_affected = total_households,
    
    # Kava
    kava_households_damaged = kava_households * get_food_security_damage_multiplier(Intensity, "Cash Crop Households", "kava", Region),
    kava_production_damaged = kava_production * get_food_security_damage_multiplier(Intensity, "Cash Crop Production", "kava", Region),
    
    # Coconut
    coconut_households_damaged = coconut_households * get_food_security_damage_multiplier(Intensity, "Cash Crop Households", "coconut", Region),
    coconut_production_damaged = coconut_production * get_food_security_damage_multiplier(Intensity, "Cash Crop Production", "coconut", Region),
    
    # Cocoa
    cocoa_households_damaged = cocoa_households * get_food_security_damage_multiplier(Intensity, "Cash Crop Households", "cocoa", Region),
    cocoa_production_damaged = cocoa_production * get_food_security_damage_multiplier(Intensity, "Cash Crop Production", "cocoa", Region),
    
    # Coffee
    coffee_households_damaged = coffee_households * get_food_security_damage_multiplier(Intensity, "Cash Crop Households", "coffee", Region),
    coffee_production_damaged = coffee_production * get_food_security_damage_multiplier(Intensity, "Cash Crop Production", "coffee", Region),
    
    # Vanilla
    vanilla_households_damaged = vanilla_households * get_food_security_damage_multiplier(Intensity, "Cash Crop Households", "vanilla", Region),
    vanilla_production_damaged = vanilla_production * get_food_security_damage_multiplier(Intensity, "Cash Crop Production", "vanilla", Region),
    
    # Tahitian Lime
    `tahitian lime_households_damaged` = `tahitian lime_households` * get_food_security_damage_multiplier(Intensity, "Cash Crop Households", "tahitian lime", Region),
    `tahitian lime_production_damaged` = `tahitian lime_production` * get_food_security_damage_multiplier(Intensity, "Cash Crop Production", "tahitian lime", Region),
    
    # Pepper
    pepper_households_damaged = pepper_households * get_food_security_damage_multiplier(Intensity, "Cash Crop Households", "pepper", Region),
    pepper_production_damaged = pepper_production * get_food_security_damage_multiplier(Intensity, "Cash Crop Production", "pepper", Region),
    
    # Noni
    noni_households_damaged = noni_households * get_food_security_damage_multiplier(Intensity, "Cash Crop Households", "noni", Region),
    noni_production_damaged = noni_production * get_food_security_damage_multiplier(Intensity, "Cash Crop Production", "noni", Region)
  ) %>%
  ungroup() %>%
  select(Region, total_households_affected, contains("_damaged"))

# Aggregate to province and national levels
cash_damage_full <- compute_council_aggregates(cash_damage)

# Reorder columns according to client's specification
cash_damage_ordered <- cash_damage_full %>%
  select(
    Region,
    total_households_affected,
    # Kava
    kava_households_damaged,
    kava_production_damaged,
    # Coconut
    coconut_households_damaged,
    coconut_production_damaged,
    # Cocoa
    cocoa_households_damaged,
    cocoa_production_damaged,
    # Coffee
    coffee_households_damaged,
    coffee_production_damaged,
    # Vanilla
    vanilla_households_damaged,
    vanilla_production_damaged,
    # Tahitian Lime
    `tahitian lime_households_damaged`,
    `tahitian lime_production_damaged`,
    # Pepper
    pepper_households_damaged,
    pepper_production_damaged,
    # Noni
    noni_households_damaged,
    noni_production_damaged
  )

# === EXPORT TABLE 2 TO CSV ===
write.csv(
  cash_damage_ordered,
  here::here("output", "FoodSecurity_02b_damage_cash_crops.csv"),
  row.names = FALSE
)

# === PRESENTATION TABLE 2 ===
formatted2 <- format_table(cash_damage_ordered)

reactable(
  formatted2,
  columns = list(
    Region = colDef(
      name = "Region", 
      minWidth = 150,
      sticky = "left",
      html = TRUE
    ),
    total_households_affected = colDef(name = "Total", format = colFormat(digits = 0)),
    
    # Kava
    kava_households_damaged = colDef(name = "Households", format = colFormat(digits = 0)),
    kava_production_damaged = colDef(name = "Production", format = colFormat(digits = 0)),
    
    # Coconut
    coconut_households_damaged = colDef(name = "Households", format = colFormat(digits = 0)),
    coconut_production_damaged = colDef(name = "Production", format = colFormat(digits = 0)),
    
    # Cocoa
    cocoa_households_damaged = colDef(name = "Households", format = colFormat(digits = 0)),
    cocoa_production_damaged = colDef(name = "Production", format = colFormat(digits = 0)),
    
    # Coffee
    coffee_households_damaged = colDef(name = "Households", format = colFormat(digits = 0)),
    coffee_production_damaged = colDef(name = "Production", format = colFormat(digits = 0)),
    
    # Vanilla
    vanilla_households_damaged = colDef(name = "Households", format = colFormat(digits = 0)),
    vanilla_production_damaged = colDef(name = "Production", format = colFormat(digits = 0)),
    
    # Tahitian Lime
    `tahitian lime_households_damaged` = colDef(name = "Households", format = colFormat(digits = 0)),
    `tahitian lime_production_damaged` = colDef(name = "Production", format = colFormat(digits = 0)),
    
    # Pepper
    pepper_households_damaged = colDef(name = "Households", format = colFormat(digits = 0)),
    pepper_production_damaged = colDef(name = "Production", format = colFormat(digits = 0)),
    
    # Noni
    noni_households_damaged = colDef(name = "Households", format = colFormat(digits = 0)),
    noni_production_damaged = colDef(name = "Production", format = colFormat(digits = 0))
  ),
  columnGroups = list(
    colGroup(name = "Total Households", columns = c("total_households_affected")),
    colGroup(name = "Kava", columns = c("kava_households_damaged", "kava_production_damaged")),
    colGroup(name = "Coconut", columns = c("coconut_households_damaged", "coconut_production_damaged")),
    colGroup(name = "Cocoa", columns = c("cocoa_households_damaged", "cocoa_production_damaged")),
    colGroup(name = "Coffee", columns = c("coffee_households_damaged", "coffee_production_damaged")),
    colGroup(name = "Vanilla", columns = c("vanilla_households_damaged", "vanilla_production_damaged")),
    colGroup(name = "Tahitian Lime", columns = c("tahitian lime_households_damaged", "tahitian lime_production_damaged")),
    colGroup(name = "Pepper", columns = c("pepper_households_damaged", "pepper_production_damaged")),
    colGroup(name = "Noni", columns = c("noni_households_damaged", "noni_production_damaged"))
  ),
  striped = TRUE,
  highlight = TRUE,
  bordered = TRUE,
  theme = reactableTheme(
    headerStyle = list(
      "&:hover" = list(background = "#eee")
    ),
    borderColor = "#ddd",
    stripedColor = "#f6f8fa"
  ),
  defaultPageSize = 15,
  showPageSizeOptions = TRUE,
  pageSizeOptions = c(10, 15, 25, 50, 75, 100),
  showSortable = TRUE
)
```


## Immediate Response Resources

```{r}
## Resources needed for Food Security restoration

# === Staple Crops Resources ===

# Helper function to get resource multipliers
get_food_security_resource_multiplier <- function(resource_type, area_council) {
  multiplier <- resource_config %>%
    filter(
      Cluster == "Food Security",
      Indicator == resource_type,
      `Area Council` == area_council
    ) %>%
    pull(Value)
  
  return(ifelse(length(multiplier) > 0, multiplier, 0))
}

# Calculate resources needed based on total affected households
resources_needed <- staple_damage %>%
  rowwise() %>%
  mutate(
    # Number of affected households (for the first column)
    num_households_affected = total_households_affected,
    
    # Resources = total affected households * resource multiplier
    island_cabbage_cuttings = round(total_households_affected * get_food_security_resource_multiplier("Island Cabbage Cuttings", Region), 0),
    taro_seedlings = round(total_households_affected * get_food_security_resource_multiplier("Taro Seedlings", Region), 0),
    kumala_cuttings = round(total_households_affected * get_food_security_resource_multiplier("Kumala Cuttings", Region), 0),
    manioc_cuttings = round(total_households_affected * get_food_security_resource_multiplier("Manioc Cuttings", Region), 0),
    yam_cuttings = round(total_households_affected * get_food_security_resource_multiplier("Yam Cuttings", Region), 0)
  ) %>%
  ungroup() %>%
  select(Region, num_households_affected, island_cabbage_cuttings, taro_seedlings, kumala_cuttings, manioc_cuttings, yam_cuttings)

# Aggregate to province and national levels
resources_needed_full <- compute_council_aggregates(resources_needed)

# === EXPORT TO CSV ===
write.csv(
  resources_needed_full,
  here::here("output", "FoodSecurity_03_resources_needed.csv"),
  row.names = FALSE
)

# === PRESENTATION ===
formatted <- format_table(resources_needed_full)

reactable(
  formatted,
  columns = list(
    Region = colDef(
      name = "Region", 
      minWidth = 150,
      sticky = "left",
      html = TRUE
    ),
    num_households_affected = colDef(
      name = "Number of Households", 
      format = colFormat(digits = 0)
    ),
    island_cabbage_cuttings = colDef(
      name = "Island Cabbage Cuttings", 
      format = colFormat(digits = 0)
    ),
    taro_seedlings = colDef(
      name = "Taro Seedlings", 
      format = colFormat(digits = 0)
    ),
    kumala_cuttings = colDef(
      name = "Kumala Cuttings", 
      format = colFormat(digits = 0)
    ),
    manioc_cuttings = colDef(
      name = "Manioc Cuttings", 
      format = colFormat(digits = 0)
    ),
    yam_cuttings = colDef(
      name = "Yam Cuttings", 
      format = colFormat(digits = 0)
    )
  ),
  columnGroups = list(
    colGroup(name = "Region", columns = c("Region")),
    colGroup(name = "Staple Crops Resources", columns = c("num_households_affected", "island_cabbage_cuttings", "taro_seedlings", "kumala_cuttings", "manioc_cuttings", "yam_cuttings"))
  ),
  striped = TRUE,
  highlight = TRUE,
  bordered = TRUE,
  theme = reactableTheme(
    headerStyle = list(
      "&:hover" = list(background = "#eee")
    ),
    borderColor = "#ddd",
    stripedColor = "#f6f8fa"
  ),
  defaultPageSize = 15,
  showPageSizeOptions = TRUE,
  pageSizeOptions = c(10, 15, 25, 50, 75, 100),
  showSortable = TRUE
)
```


# Estimated Financial Damage 

### Table 1: Staple Crops Financial Damage

```{r}
# Helper function to get unit costs for food security
get_food_security_unit_cost <- function(indicator_type, crop_type, area_council) {
  unit_cost <- financial_config %>%
    filter(
      Cluster == "Food Security",
      Indicator == indicator_type,
      tolower(trimws(Attribute)) == tolower(trimws(crop_type)),
      `Area Council` == area_council
    ) %>%
    pull(Value)
  
  return(ifelse(length(unit_cost) > 0, unit_cost, 0))
}

# Calculate staple crops financial damage
staple_financial <- staple_damage %>%
  rowwise() %>%
  mutate(
    # Financial damage for each crop (damaged production * unit cost)
    island_cabbage_financial = `island cabbage_production_damaged` * get_food_security_unit_cost("Staple Crop Production", "island cabbage", Region),
    banana_financial = banana_production_damaged * get_food_security_unit_cost("Staple Crop Production", "banana", Region),
    taro_financial = taro_production_damaged * get_food_security_unit_cost("Staple Crop Production", "taro", Region),
    kumala_financial = kumala_production_damaged * get_food_security_unit_cost("Staple Crop Production", "kumala", Region),
    manioc_financial = manioc_production_damaged * get_food_security_unit_cost("Staple Crop Production", "manioc", Region),
    yam_financial = yam_production_damaged * get_food_security_unit_cost("Staple Crop Production", "yam", Region),
    
    # Total staple crops financial damage
    total_staple_financial = island_cabbage_financial + banana_financial + taro_financial + 
                             kumala_financial + manioc_financial + yam_financial
  ) %>%
  ungroup() %>%
  select(Region, total_staple_financial, island_cabbage_financial, banana_financial, 
         taro_financial, kumala_financial, manioc_financial, yam_financial)

# Aggregate to province and national levels
staple_financial_full <- compute_council_aggregates(staple_financial)

# Reorder columns
staple_financial_ordered <- staple_financial_full %>%
  select(Region, total_staple_financial, island_cabbage_financial, banana_financial, 
         taro_financial, kumala_financial, manioc_financial, yam_financial)

# === EXPORT TABLE 1 TO CSV ===
write.csv(
  staple_financial_ordered,
  here::here("output", "FoodSecurity_04a_financial_damage_staple_crops.csv"),
  row.names = FALSE
)

# === PRESENTATION TABLE 1 ===
formatted1 <- format_table(staple_financial_ordered)

reactable(
  formatted1,
  columns = list(
    Region = colDef(
      name = "Region", 
      minWidth = 150,
      sticky = "left",
      html = TRUE
    ),
    total_staple_financial = colDef(
      name = "Total Value",
      format = colFormat(suffix = " VT", separators = TRUE, digits = 0)
    ),
    island_cabbage_financial = colDef(
      name = "Island Cabbage",
      format = colFormat(suffix = " VT", separators = TRUE, digits = 0)
    ),
    banana_financial = colDef(
      name = "Banana",
      format = colFormat(suffix = " VT", separators = TRUE, digits = 0)
    ),
    taro_financial = colDef(
      name = "Taro",
      format = colFormat(suffix = " VT", separators = TRUE, digits = 0)
    ),
    kumala_financial = colDef(
      name = "Kumala",
      format = colFormat(suffix = " VT", separators = TRUE, digits = 0)
    ),
    manioc_financial = colDef(
      name = "Manioc",
      format = colFormat(suffix = " VT", separators = TRUE, digits = 0)
    ),
    yam_financial = colDef(
      name = "Yam",
      format = colFormat(suffix = " VT", separators = TRUE, digits = 0)
    )
  ),
  columnGroups = list(
    colGroup(name = "Region", columns = c("Region")),
    colGroup(name = "Staple Crops Financial Damage (VT)", 
             columns = c("total_staple_financial", "island_cabbage_financial", "banana_financial", 
                        "taro_financial", "kumala_financial", "manioc_financial", "yam_financial"))
  ),
  striped = TRUE,
  highlight = TRUE,
  bordered = TRUE,
  theme = reactableTheme(
    headerStyle = list(
      "&:hover" = list(background = "#eee")
    ),
    borderColor = "#ddd",
    stripedColor = "#f6f8fa"
  ),
  defaultPageSize = 15,
  showPageSizeOptions = TRUE,
  pageSizeOptions = c(10, 15, 25, 50, 75, 100),
  showSortable = TRUE
)
```


### Table 2: Cash Crops Financial Damage


```{r}
# Calculate cash crops financial damage
cash_financial <- cash_damage %>%
  rowwise() %>%
  mutate(
    # Financial damage for each crop (damaged production * unit cost)
    kava_financial = kava_production_damaged * get_food_security_unit_cost("Cash Crop Production", "kava", Region),
    coconut_financial = coconut_production_damaged * get_food_security_unit_cost("Cash Crop Production", "coconut", Region),
    cocoa_financial = cocoa_production_damaged * get_food_security_unit_cost("Cash Crop Production", "cocoa", Region),
    coffee_financial = coffee_production_damaged * get_food_security_unit_cost("Cash Crop Production", "coffee", Region),
    vanilla_financial = vanilla_production_damaged * get_food_security_unit_cost("Cash Crop Production", "vanilla", Region),
    tahitian_lime_financial = `tahitian lime_production_damaged` * get_food_security_unit_cost("Cash Crop Production", "tahitian lime", Region),
    pepper_financial = pepper_production_damaged * get_food_security_unit_cost("Cash Crop Production", "pepper", Region),
    noni_financial = noni_production_damaged * get_food_security_unit_cost("Cash Crop Production", "noni", Region),
    
    # Total cash crops financial damage
    total_cash_financial = kava_financial + coconut_financial + cocoa_financial + 
                           coffee_financial + vanilla_financial + tahitian_lime_financial + 
                           pepper_financial + noni_financial
  ) %>%
  ungroup() %>%
  select(Region, total_cash_financial, kava_financial, coconut_financial, cocoa_financial,
         coffee_financial, vanilla_financial, tahitian_lime_financial, pepper_financial, noni_financial)

# Aggregate to province and national levels
cash_financial_full <- compute_council_aggregates(cash_financial)

# Reorder columns
cash_financial_ordered <- cash_financial_full %>%
  select(Region, total_cash_financial, kava_financial, coconut_financial, cocoa_financial,
         coffee_financial, vanilla_financial, tahitian_lime_financial, pepper_financial, noni_financial)

# === EXPORT TABLE 2 TO CSV ===
write.csv(
  cash_financial_ordered,
  here::here("output", "FoodSecurity_04b_financial_damage_cash_crops.csv"),
  row.names = FALSE
)

# === PRESENTATION TABLE 2 ===
formatted2 <- format_table(cash_financial_ordered)

reactable(
  formatted2,
  columns = list(
    Region = colDef(
      name = "Region", 
      minWidth = 150,
      sticky = "left",
      html = TRUE
    ),
    total_cash_financial = colDef(
      name = "Total Value",
      format = colFormat(suffix = " VT", separators = TRUE, digits = 0)
    ),
    kava_financial = colDef(
      name = "Kava",
      format = colFormat(suffix = " VT", separators = TRUE, digits = 0)
    ),
    coconut_financial = colDef(
      name = "Coconut",
      format = colFormat(suffix = " VT", separators = TRUE, digits = 0)
    ),
    cocoa_financial = colDef(
      name = "Cocoa",
      format = colFormat(suffix = " VT", separators = TRUE, digits = 0)
    ),
    coffee_financial = colDef(
      name = "Coffee",
      format = colFormat(suffix = " VT", separators = TRUE, digits = 0)
    ),
    vanilla_financial = colDef(
      name = "Vanilla",
      format = colFormat(suffix = " VT", separators = TRUE, digits = 0)
    ),
    tahitian_lime_financial = colDef(
      name = "Tahitian Lime",
      format = colFormat(suffix = " VT", separators = TRUE, digits = 0)
    ),
    pepper_financial = colDef(
      name = "Pepper",
      format = colFormat(suffix = " VT", separators = TRUE, digits = 0)
    ),
    noni_financial = colDef(
      name = "Noni",
      format = colFormat(suffix = " VT", separators = TRUE, digits = 0)
    )
  ),
  columnGroups = list(
    colGroup(name = "Region", columns = c("Region")),
    colGroup(name = "Cash Crops Financial Damage (VT)", 
             columns = c("total_cash_financial", "kava_financial", "coconut_financial", "cocoa_financial",
                        "coffee_financial", "vanilla_financial", "tahitian_lime_financial", 
                        "pepper_financial", "noni_financial"))
  ),
  striped = TRUE,
  highlight = TRUE,
  bordered = TRUE,
  theme = reactableTheme(
    headerStyle = list(
      "&:hover" = list(background = "#eee")
    ),
    borderColor = "#ddd",
    stripedColor = "#f6f8fa"
  ),
  defaultPageSize = 15,
  showPageSizeOptions = TRUE,
  pageSizeOptions = c(10, 15, 25, 50, 75, 100),
  showSortable = TRUE
)
```





```{r}

```
