---
title: "Disaster Damage and Response Estimation"
author: "Théophile L. Mouton"
date: "`r Sys.Date()`"
include-after-body: "asset/footer.html"
css: "asset/style.css"
format:
  html:
    template-partials: 
      - asset/title-block.html
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    highlight-style: github
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    theme: cosmo
    fig-width: 8
    fig-height: 6
    fig-dpi: 300
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
  error: false
---

# Intro

This file runs based on a file called `config.csv`. This file provides the cyclone category applied to each council in Vanuatu. It has just 2 columns:
- council name
- cyclone category (number from 3 to 5)

Let's load this file to use it later in our analysis:

```{r}
config <- read.csv("data/config.csv")
```

Another file that will be useful: a 2 columns dataframe that provides the list of councils with their Province
```{r}
council_province_lookup <- read.csv("data/council_province_lookup.csv")
```


And load the required libs
```{r, warning=FALSE, message=FALSE}
# Load required packages
library(dplyr)
library(tidyr)
library(reactable)
library(htmltools)
library(readxl)
library(here)
```


Order of councils and provinces in the final tables:

```{r}
# Define the custom order for regions
region_order <- c(
  "National", # Will be displayed as Vanuatu
  "Torba", "Torres", "Ureparapara", "Motalava", "West Vanualava", "East Vanualava", 
  "Mota", "East Gaua", "West Gaua", "Merelava",
  "Sanma", "Luganville", "North West Santo", "Big Bay Coast", "Big Bay Inland", 
  "West Santo", "South Santo 1", "South Santo 2", "East Santo", "South East Santo", 
  "Canal - Fanafo", "West Malo", "East Malo",
  "Penama", "West Ambae", "North Ambae", "East Ambae", "South Ambae", "North Maewo", 
  "South Maewo", "North Pentecost", "Central Pentecost 1", "Central Pentecost 2", 
  "South Pentecost",
  "Malampa", "North West Malekula", "North East Malekula", "Central Malekula", 
  "South West Malekula", "South East Malekula", "South Malekula", "North Ambrym", 
  "West Ambrym", "South East Ambrym", "Paama",
  "Shefa", "Port Vila", "Vermali", "Vermaul", "Varisu", "South Epi", "North Tongoa", 
  "Tongariki", "Makimae", "Nguna", "Emau", "Malorua", "North Efate", "Mele", 
  "Tanvasoko", "Ifira", "Pango", "Erakor", "Eratap", "Eton",
  "Tafea", "North Erromango", "South Erromango", "Aniwa", "North Tanna", "West Tanna", 
  "Middle Bush Tanna", "South West Tanna", "Whitesands", "South Tanna", "Futuna", "Aneityum"
)

# Provincial levels (to be bolded)
provinces <- c("National", "Torba", "Sanma", "Penama", "Malampa", "Shefa", "Tafea")
```

A function to aggregate a table with one row per council. It adds value per province and for the whole country
```{r}
# council_data must be a data frame with a column called "Region"
compute_council_aggregates <- function(council_data) {
  
  province_data <- council_data %>%
    left_join(council_province_lookup, by = c("Region" = "Council")) %>%
    group_by(Province) %>%
    summarise(across(where(is.numeric), sum, na.rm = TRUE), .groups = "drop") %>%
    rename(Region = Province)  
  
  national_data <- province_data %>%
    summarise(across(where(is.numeric), sum, na.rm = TRUE)) %>%
    mutate(Region = "National")
  
  result <- bind_rows(
    national_data,
    province_data,
    council_data
  )
  
  # Add default order and formatting columns
  result <- result %>%
    mutate(default_order = match(Region, region_order)) %>%
    arrange(default_order) %>%
    select(-default_order)

  return(result)
}

format_table <- function(data){
  result <- data %>%
    select(Region, everything()) %>% 
    mutate(
      Region = ifelse(
        Region %in% provinces, 
        paste0("<b>", Region, "</b>"), 
        Region
      )
    )
  return(result)
}
```







# Education

## Baseline: Number of Schools, Students, and Teachers

```{r}
# Read the Excel file
full_data <- read_excel(here("data", "11b. Full Data Set.xlsx"))

# === DATA WRANGLING ===
# Filter for Education cluster and Area Council level data
education_data <- full_data %>%
  filter(Cluster == "Education") %>%
  filter(!is.na(`Area Council`)) %>%
  mutate(Region = `Area Council`)

# Reshape the data to get Schools, Students, Teachers as separate columns
education_wide <- education_data %>%
  mutate(
    education_level = tolower(Attribute),  # Keep lowercase for consistency
    measure_type = case_when(
      grepl("Number Schools", Indicator) ~ "schools",
      grepl("Students", Indicator) ~ "students",
      grepl("Teachers", Indicator) ~ "teachers",
      TRUE ~ "other"
    )
  ) %>%
  filter(measure_type %in% c("schools", "students", "teachers")) %>%
  select(Region, education_level, measure_type, Value) %>%
  pivot_wider(
    names_from = c(education_level, measure_type),
    values_from = Value,
    names_sep = "_"
  ) %>%
  mutate(across(where(is.numeric), ~replace_na(.x, 0)))

# Compute aggregates (province and national levels)
education_aggregated <- compute_council_aggregates(education_wide)

# === PRESENTATION ===
# Format for display
formatted <- format_table(education_aggregated)

# Create the reactable with proper case labels
reactable(
  formatted,
  columns = list(
    Region = colDef(
      name = "Region", 
      minWidth = 150,
      sortable = TRUE,
      defaultSortOrder = "asc",
      html = TRUE,
      sticky = "left"
    ),
    
    # ECCE columns
    ecce_schools = colDef(name = "Schools", format = colFormat(digits = 0)),
    ecce_students = colDef(name = "Students", format = colFormat(digits = 0)),
    ecce_teachers = colDef(name = "Teachers", format = colFormat(digits = 0)),
    
    # Primary columns  
    primary_schools = colDef(name = "Schools", format = colFormat(digits = 0)),
    primary_students = colDef(name = "Students", format = colFormat(digits = 0)),
    primary_teachers = colDef(name = "Teachers", format = colFormat(digits = 0)),
    
    # Secondary columns
    secondary_schools = colDef(name = "Schools", format = colFormat(digits = 0)),
    secondary_students = colDef(name = "Students", format = colFormat(digits = 0)),
    secondary_teachers = colDef(name = "Teachers", format = colFormat(digits = 0))
  ),
  columnGroups = list(
    colGroup(name = "Region", columns = c("Region")),
    colGroup(name = "ECCE", columns = c("ecce_schools", "ecce_students", "ecce_teachers")),
    colGroup(name = "Primary", columns = c("primary_schools", "primary_students", "primary_teachers")),
    colGroup(name = "Secondary", columns = c("secondary_schools", "secondary_students", "secondary_teachers"))
  ),
  striped = TRUE,
  highlight = TRUE,
  bordered = TRUE,
  theme = reactableTheme(
    headerStyle = list(
      "&:hover" = list(background = "#eee")
    ),
    borderColor = "#ddd",
    stripedColor = "#f6f8fa"
  ),
  defaultPageSize = 15,
  showPageSizeOptions = TRUE,
  pageSizeOptions = c(10, 15, 25, 50, 75, 100),
  showSortable = TRUE
)
```





## Estimating damage: Number of damaged schools and students affected

```{r}
# === DATA WRANGLING ===
# Add cyclone strength to baseline values
education_with_config <- education_wide %>%
  left_join(config, by = c("Region" = "council"))

# Load the baseline multiplying factors
baseline_factors <- read_excel(here("data", "12. Baseline_Multiplying_Factors.xlsx"))

# Helper function for damage multipliers
get_damage_multiplier <- function(cyclone_category, education_level, measure_type) {
  multiplier <- baseline_factors %>%
    filter(
      Cluster == "Education",
      Attribute == education_level,  # Now both are lowercase
      grepl(measure_type, Indicator, ignore.case = TRUE)
    ) %>%
    pull(!!paste0("Category ", cyclone_category))
  
  return(ifelse(length(multiplier) > 0, multiplier, 0))
}

# Calculate damage estimates
damage_estimates <- education_with_config %>%
  filter(!is.na(category)) %>%
  rowwise() %>%
  mutate(
    # Changed from ECCE_Schools to ecce_schools (lowercase)
    ecce_schools_damaged = round(ecce_schools * get_damage_multiplier(category, "ecce", "schools")),
    ecce_students_affected = round(ecce_students * get_damage_multiplier(category, "ecce", "students")),
    
    # Changed from Primary_Schools to primary_schools (lowercase)
    primary_schools_damaged = round(primary_schools * get_damage_multiplier(category, "primary", "schools")),
    primary_students_affected = round(primary_students * get_damage_multiplier(category, "primary", "students")),
    
    # Changed from Secondary_Schools to secondary_schools (lowercase)
    secondary_schools_damaged = round(secondary_schools * get_damage_multiplier(category, "secondary", "schools")),
    secondary_students_affected = round(secondary_students * get_damage_multiplier(category, "secondary", "students"))
  ) %>%
  ungroup() %>%
  select(Region, contains("_damaged"), contains("_affected"))

# Aggregate to province and national levels
damage_estimates_full <- compute_council_aggregates(damage_estimates)

# === PRESENTATION ===
# Format for display
formatted <- format_table(damage_estimates_full)

# Create the damage estimation reactable
reactable(
  formatted,
  columns = list(
    Region = colDef(
      name = "Region", 
      minWidth = 150,
      sticky = "left",
      html = TRUE
    ),
    
    # ECCE columns
    ecce_schools_damaged = colDef(name = "Schools Damaged", format = colFormat(digits = 0)),
    ecce_students_affected = colDef(name = "Students Affected", format = colFormat(digits = 0)),
    
    # Primary columns  
    primary_schools_damaged = colDef(name = "Schools Damaged", format = colFormat(digits = 0)),
    primary_students_affected = colDef(name = "Students Affected", format = colFormat(digits = 0)),
    
    # Secondary columns
    secondary_schools_damaged = colDef(name = "Schools Damaged", format = colFormat(digits = 0)),
    secondary_students_affected = colDef(name = "Students Affected", format = colFormat(digits = 0))
  ),
  columnGroups = list(
    colGroup(name = "Region", columns = c("Region")),
    colGroup(name = "ECCE", columns = c("ecce_schools_damaged", "ecce_students_affected")),
    colGroup(name = "Primary", columns = c("primary_schools_damaged", "primary_students_affected")),
    colGroup(name = "Secondary", columns = c("secondary_schools_damaged", "secondary_students_affected"))
  ),
  striped = TRUE,
  highlight = TRUE,
  bordered = TRUE,
  theme = reactableTheme(
    headerStyle = list(
      "&:hover" = list(background = "#eee")
    ),
    borderColor = "#ddd",
    stripedColor = "#f6f8fa"
  ),
  defaultPageSize = 15,
  showPageSizeOptions = TRUE,
  pageSizeOptions = c(10, 15, 25, 50, 75, 100),
  showSortable = TRUE
)

```






## Resources needed to be sent to those affected

```{r}
# Load the resource response configuration
resource_config <- read_excel(here("data", "13. Resource_Sent_Response.xlsx"))

# Create a function to get resource multipliers from config
get_resource_multiplier <- function(resource_type) {
  multiplier <- resource_config %>%
    filter(Cluster == "Education", Indicator == resource_type) %>%
    pull(Value)
  
  return(ifelse(length(multiplier) > 0, multiplier, 0))
}

# Configuration parameters
days_of_support <- 14  # Number of days to provide resources
default_cyclone_category <- 4  # Default cyclone category (can be customized)

# Function to calculate resources needed
calculate_resources <- function(region_data, cyclone_category, days = 14) {
  
  region_data %>%
    mutate(
      # Get damage multiplier for this cyclone category
      damage_multiplier = get_damage_multiplier(cyclone_category, "ecce", "Schools"), # Same multiplier for all
      
      # ECCE Resources
      ECCE_Tents = round(ECCE_Schools * damage_multiplier * get_resource_multiplier("Tent"), 0),
      ECCE_Solar_Lamps = round(ECCE_Schools * damage_multiplier * get_resource_multiplier("Solar Lamp"), 0),
      ECCE_Water = ceiling(ECCE_Students * damage_multiplier * get_resource_multiplier("Water") * days), # Round up for bottles
      ECCE_Rice = round(ECCE_Students * damage_multiplier * get_resource_multiplier("Rice") * days, 1),
      ECCE_Tin_Fish = round(ECCE_Students * damage_multiplier * get_resource_multiplier("Tin fish") * days, 0),
      
      # Primary Resources  
      Primary_Tents = round(Primary_Schools * damage_multiplier * get_resource_multiplier("Tent"), 0),
      Primary_Solar_Lamps = round(Primary_Schools * damage_multiplier * get_resource_multiplier("Solar Lamp"), 0),
      Primary_Water = ceiling(Primary_Students * damage_multiplier * get_resource_multiplier("Water") * days), # Round up for bottles
      Primary_Rice = round(Primary_Students * damage_multiplier * get_resource_multiplier("Rice") * days, 1),
      Primary_Tin_Fish = round(Primary_Students * damage_multiplier * get_resource_multiplier("Tin fish") * days, 0),
      
      # Secondary Resources
      Secondary_Tents = round(Secondary_Schools * damage_multiplier * get_resource_multiplier("Tent"), 0),
      Secondary_Solar_Lamps = round(Secondary_Schools * damage_multiplier * get_resource_multiplier("Solar Lamp"), 0),
      Secondary_Water = ceiling(Secondary_Students * damage_multiplier * get_resource_multiplier("Water") * days), # Round up for bottles
      Secondary_Rice = round(Secondary_Students * damage_multiplier * get_resource_multiplier("Rice") * days, 1),
      Secondary_Tin_Fish = round(Secondary_Students * damage_multiplier * get_resource_multiplier("Tin fish") * days, 0),
      
      # Total Resources (sum across all education levels)
      Total_Tents = ECCE_Tents + Primary_Tents + Secondary_Tents,
      Total_Solar_Lamps = ECCE_Solar_Lamps + Primary_Solar_Lamps + Secondary_Solar_Lamps,
      Total_Water = ECCE_Water + Primary_Water + Secondary_Water, # Already whole numbers
      Total_Rice = round(ECCE_Rice + Primary_Rice + Secondary_Rice, 1),
      Total_Tin_Fish = ECCE_Tin_Fish + Primary_Tin_Fish + Secondary_Tin_Fish
    ) %>%
    select(Region, contains("Tents"), contains("Solar_Lamps"), contains("Water"), contains("Rice"), contains("Tin_Fish"))
}

# Apply resource calculations to all regions using the education_wide dataset
resources_needed <- education_wide %>%
  calculate_resources(default_cyclone_category, days_of_support)

resources_needed_full <- compute_council_aggregates(resources_needed)
formatted <- format_table(resources_needed_full)

# Create the resources needed reactable
reactable(
  formatted,
  columns = list(
    Region = colDef(
      name = "Region", 
      minWidth = 150,
      sticky = "left",
      html = TRUE # Allow HTML formatting for bold text
    ),
    
    # ECCE columns
    ECCE_Tents = colDef(name = "Tents", format = colFormat(digits = 0)),
    ECCE_Solar_Lamps = colDef(name = "Solar Lamps", format = colFormat(digits = 0)),
    ECCE_Water = colDef(name = "Water (1L bottles)", format = colFormat(digits = 0)),
    ECCE_Rice = colDef(name = "Rice (kg)", format = colFormat(digits = 1)),
    ECCE_Tin_Fish = colDef(name = "Tinned Fish", format = colFormat(digits = 0)),
    
    # Primary columns  
    Primary_Tents = colDef(name = "Tents", format = colFormat(digits = 0)),
    Primary_Solar_Lamps = colDef(name = "Solar Lamps", format = colFormat(digits = 0)),
    Primary_Water = colDef(name = "Water (1L bottles)", format = colFormat(digits = 0)),
    Primary_Rice = colDef(name = "Rice (kg)", format = colFormat(digits = 1)),
    Primary_Tin_Fish = colDef(name = "Tinned Fish", format = colFormat(digits = 0)),
    
    # Secondary columns
    Secondary_Tents = colDef(name = "Tents", format = colFormat(digits = 0)),
    Secondary_Solar_Lamps = colDef(name = "Solar Lamps", format = colFormat(digits = 0)),
    Secondary_Water = colDef(name = "Water (1L bottles)", format = colFormat(digits = 0)),
    Secondary_Rice = colDef(name = "Rice (kg)", format = colFormat(digits = 1)),
    Secondary_Tin_Fish = colDef(name = "Tinned Fish", format = colFormat(digits = 0)),
    
    # Total columns
    Total_Tents = colDef(name = "Tents", format = colFormat(digits = 0)),
    Total_Solar_Lamps = colDef(name = "Solar Lamps", format = colFormat(digits = 0)),
    Total_Water = colDef(name = "Water (1L bottles)", format = colFormat(digits = 0)),
    Total_Rice = colDef(name = "Rice (kg)", format = colFormat(digits = 1)),
    Total_Tin_Fish = colDef(name = "Tinned Fish", format = colFormat(digits = 0))
  ),
  columnGroups = list(
    colGroup(name = "Region", columns = c("Region")),
    colGroup(name = "ECCE", columns = c("ECCE_Tents", "ECCE_Solar_Lamps", "ECCE_Water", "ECCE_Rice", "ECCE_Tin_Fish")),
    colGroup(name = "Primary", columns = c("Primary_Tents", "Primary_Solar_Lamps", "Primary_Water", "Primary_Rice", "Primary_Tin_Fish")),
    colGroup(name = "Secondary", columns = c("Secondary_Tents", "Secondary_Solar_Lamps", "Secondary_Water", "Secondary_Rice", "Secondary_Tin_Fish")),
    colGroup(name = "Totals", columns = c("Total_Tents", "Total_Solar_Lamps", "Total_Water", "Total_Rice", "Total_Tin_Fish"))
  ),
  striped = TRUE,
  highlight = TRUE,
  bordered = TRUE,
  theme = reactableTheme(
    headerStyle = list(
      "&:hover" = list(background = "#eee")
    ),
    borderColor = "#ddd",
    stripedColor = "#f6f8fa"
  ),
  defaultPageSize = 15,
  showPageSizeOptions = TRUE,
  pageSizeOptions = c(10, 15, 25, 50, 75, 100),
  showSortable = TRUE
)
```




## Estimate financial damage from cyclone

```{r, eval=FALSE}
# Load the financial damage estimation config
financial_config <- read_excel(here("data", "14. Estimation_Financial_Damage.xlsx"))

# Clean column names to remove any extra spaces
names(financial_config) <- trimws(names(financial_config))

# Create a function to get unit costs from config
get_unit_cost <- function(school_type) {
  unit_cost <- financial_config %>%
    filter(trimws(Cluster) == "Education", 
           trimws(tolower(Attribute)) == tolower(school_type),
           grepl("Number Schools", trimws(Indicator), ignore.case = TRUE)) %>%
    pull(Value)
  
  return(ifelse(length(unit_cost) > 0, unit_cost, 0))
}

# Configuration parameters
default_cyclone_category <- 4  # Default cyclone category (can be customized)

# Function to calculate financial damage
calculate_financial_damage <- function(region_data, cyclone_category) {
  
  region_data %>%
    mutate(
      # Get damage multiplier for this cyclone category
      damage_multiplier = get_damage_multiplier(cyclone_category, "ecce", "Schools"), # Same multiplier for all
      
      # Calculate financial damage per education level (in VT)
      ECCE_Financial_Damage = ECCE_Schools * get_unit_cost("ecce") * damage_multiplier,
      Primary_Financial_Damage = Primary_Schools * get_unit_cost("primary") * damage_multiplier,
      Secondary_Financial_Damage = Secondary_Schools * get_unit_cost("secondary") * damage_multiplier,
      
      # Calculate total financial damage
      Total_Financial_Damage = ECCE_Financial_Damage + Primary_Financial_Damage + Secondary_Financial_Damage
    ) %>%
    select(Region_Formatted, contains("Financial_Damage"))
}

# Apply financial damage calculations to all regions using the education_wide dataset
financial_damage <- education_wide %>%
  select(Region_Formatted, contains("ECCE"), contains("Primary"), contains("Secondary")) %>%
  calculate_financial_damage(default_cyclone_category)

# Create the financial damage reactable
reactable(
  financial_damage,
  columns = list(
    Region_Formatted = colDef(
      name = "Region", 
      minWidth = 150,
      sticky = "left",
      html = TRUE # Allow HTML formatting for bold text
    ),
    
    # Financial damage columns
    ECCE_Financial_Damage = colDef(
      name = "ECCE", 
      format = colFormat(suffix = " VT", separators = TRUE, digits = 0)
    ),
    Primary_Financial_Damage = colDef(
      name = "Primary", 
      format = colFormat(suffix = " VT", separators = TRUE, digits = 0)
    ),
    Secondary_Financial_Damage = colDef(
      name = "Secondary", 
      format = colFormat(suffix = " VT", separators = TRUE, digits = 0)
    ),
    Total_Financial_Damage = colDef(
      name = "Total", 
      format = colFormat(suffix = " VT", separators = TRUE, digits = 0)
    )
  ),
  striped = TRUE,
  highlight = TRUE,
  bordered = TRUE,
  theme = reactableTheme(
    headerStyle = list(
      "&:hover" = list(background = "#eee")
    ),
    borderColor = "#ddd",
    stripedColor = "#f6f8fa"
  ),
  defaultPageSize = 15,
  showPageSizeOptions = TRUE,
  pageSizeOptions = c(10, 15, 25, 50, 75, 100),
  showSortable = TRUE
)


```
